---
bibliography: references.bib
---

# A Parsimonious Computational Model Linking Appraisal and Subjective Feeling {#computational-model}

```{r computational-model-setup, include=FALSE, echo=FALSE}
library(tidyverse)
library(papaja)
library(see)
library(RColorBrewer)  
library(here)
library(knitr)
library(kableExtra)

# Load the relevant data and graphs
source(here("data", "dew-rt", "dew-rt-export.R"), local = knitr::knit_global(), encoding = "UTF-8")
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
```

This chapter illustrates the parsimonious computational model that is at the core of the Dynamic Emotion Wheel. The model is based upon the probabilistic link, stated by appraisal theories of emotions and the Component Process Model more specifically, between the cognitive evaluation of the situation and the resulting holistic experience, the subjective feeling, that can be lexicalized in natural language words or idioms. The combination of the computational model and the interface of the DEW illustrated in the previous chapter makes it possible to abstract from two-dimensional spaces and take advantage of underlying structure for which the number of dimensions can be defined according to researchers and practitioners' choice. In this regard, the chapter starts by briefly outlining the interest for affective spaces with a different number of underlying dimensions. Second, it illustrates the logic behind the computational model inside the DEW. The following three sections, then, describe how the same computational logic can be applied to one-dimensional, two-dimensional, and multi-dimensional affective spaces (3 and beyond). The chapter ends with an assessment of the limitations of the computational model.

## Interest For *N*-Dimensional Affective Spaces

As the use of the EATMINT circumplex depicted in Section \@ref(eatmint-circumplex) has shown, it is possible to adapt the underlying *reference frame* that link the appraisal dimensions with the subjective feelings. Still, the example remained in the limits of a bi-dimensional affective space. The use of this kind of affective space, though, was driven mainly by the familiarity and widespread use of the graphical representation in the literature as well as in other self-report tools, in particular the Geneva Emotion Wheel. Beyond the theoretical foundation of bi-dimensional affective spaces, the representation also happens to be the highest number of dimensions that can be easily represented graphically on a screen or sheet of paper. Nevertheless, once the graphical user interface of the Dynamic Emotion Wheel breaks the *overlapping* between the affective space and the self-report tool, than a circumplex becomes one *reference frame* among others. In other words, the circumplex determines only the underlying theoretical and/or empirically validated affective space, but does not force the interface to change dramatically. In fact, the appraisal dimensions in the DEW are *flattened* into the rating of the sliders. Concretely, more appraisal dimensions would simply translate in more sliders through which respondents rate the situation. Furthermore, the computational model can subset only a limited number of subjective feelings, which does not require thus that all of them are represented at the same time. This fact has several consequences both from an *affective science* and an *emotional awareness* perspectives, which are considered hereafter.

### Affective Science's Interest in *N*-Dimensional Spaces

As it has been shown in Part I, researchers adopt very different conceptualization of emotion across and within the same family of emotion theories (see Section \@ref(theories-of-emotion)). Theories assuming that a dimensional mechanism is implicated in the structure of emotion assume different positions on the number of dimensions, ranging from a uni-dimensional structure (*positive* vs. *negative* emotions) to the multiple sequential evaluation checks of the Component Process Model listed in Section \@ref(cpm-appraisal-module).

In this regard, empirical evidence gathered in the last decade has also highlighted how two dimensions are not enough to account for the variety of emotional experiences, even though some of them have more *weight* than others in discriminating between discrete emotions [@fontaineWorldEmotionsNot2007; @gilliozMappingEmotionTerms2016; @gentschEffectsAchievementContexts2017; @fontaineComponentsEmotionalMeaning2013]. In an attempt to provide a working definition of emotion based on seven defining features, Mulligan and Scherer [-@mulliganWorkingDefinitionEmotion2012, p. 346] propose that "$x$ is an emotion *only if* [...] $x$ is triggered by at least one appraisal, and $x$ is guided by at least one appraisal". The working definition clearly states that one appraisal is necessary, but does not provide any fixed number or upper-limit.

Even when the same number of dimensions are postulated, there is no agreement on what these dimensions should be, across and within families of emotion theories. Taking two-dimensional affective spaces as an example, we can identify at least (1) the ubiquitous *Valence* x *Arousal* dimensions [@russellCircumplexModelAffect1980; @russellPanculturalAspectsHuman1983; @stanleyTwodimensionalAffectiveSpace2009] used for instance by the *emot-control* [@feidakisProvidingEmotionAwareness2014; @feidakis2013] and Mood Meter application [@brackettRULERTheoryDrivenSystemic2019; @hoffmannTeachingEmotionRegulation2020; @nathansonCreatingEmotionallyIntelligent2016]; (2) the *Valence* x *Control/Power* dimensions of the Geneva Emotion Wheel [@schererWhatAreEmotions2005; @schererGRIDMeetsWheel2013; @shumanConceptsStructuresEmotions2014]; or (3) Pekrun's *Control* x *Value* dimensions in achievement emotions [@pekrunControlValueTheoryAchievement2014; @pekrunControlValueTheoryAchievement2006].

At the same time, the number of lexicalized emotions representing the subjective feelings does not have to be bound to 20 or any other fixed number. Once again, it should be up to the researcher or practitioner to decide whether emotion differentiation between a small or large number of options is an important part of the process [@erbasRoleValenceFocus2015; @moorsAppraisalTheoriesEmotion2013]. For instance the EMORE-L tool by @molinariEMORELOutilReporting2016 comprises 8 discrete emotions, whereas the digital version of the Mood Meter application linked to the RULER approach [@brackettRULERTheoryDrivenSystemic2019; @hoffmannTeachingEmotionRegulation2020] allows respondents to choose between 100 lexicalized emotions.

To sum-up, the DEW inner functioning not only allows to break through a forced bi-dimensional affective space, but it also permits different dimensional affective spaces to coexists within more or less the same interface. This may be interesting to assess and compare the contribution of a lower vs. a higher-order of dimensions in active rating.

### Emotional Awareness's Interest in *N*-Dimensional Spaces

The DEW adopted two dimensions for usability constraints: it was considered important that expressing an emotion would not take too much time and effort. As described in Section \@ref(instrumental-awareness), though, there is a growing consensus in considering that the time and effort in producing and consuming awareness should not be evaluated in absolute terms, but in a relative trade-off, which accounts for the instrumentality of the information provided in helping learners to improve their learning processes and outcomes [@dillenbourgSymmetryPartnerModelling2016; @buderGroupAwarenessTools2011; @janssenCoordinatedComputerSupportedCollaborative2013]. The number of dimensions should therefore be a pedagogical decision, not imposed *a priori* by the interface of the tool. For instance, in the EMORE-L tool [@molinariEMORELOutilReporting2016], learners use three dimensions to rate their emotional experience: the subjective control over the situation, the subjective value attributed to the event, and the perceived level of activation. As outlined in the abstract model of emotional awareness in Section \@ref(abstract-model-of-ea), an EAT can help learners in extrapolating meaning-making by guiding their introspection and self-reflection. Appraisal criteria -- or dimensions more broadly -- may have a privileged function in this sense, and it would be limiting to have to decide between a fixed number of them for visuo-spatial constraints.

As for the *affective science* perspective illustrated above, the same logic applies to the pre-determined lexicalized emotions that are presented to learners. Depending on the accuracy and level of granularity desired in representing the subjective feeling, the underlying affective space can provide any number of options. From a computational perspective, these options would be all the more distinct if they are *scattered* in multiple dimensions. If the use of an EAT is, for instance, directed towards improving the ability of learners in discerning between similar, but not identical, emotional experiences, having a tool that provides adjacent lexicalized emotions as a function of multiple evaluation can be a means to orient learners into fine-grained differences. The subset mechanism of the DEW provides in this regard the possibility to maintain a sheer number of options, without having them all together on the interface.

A higher number of dimensions in self-reporting emotions also provide more accurate information for the graphical representation in the perceiving-monitoring function of awareness. From a reverse-engineering perspective, it is easier to answer the questions "what does my colleague mean when she says she feels $x$?", when $x$ can be quantified on different appraisal dimensions. For instance, with *Valence* alone, the learner can only infer if $x$ is roughly pleasant or unpleasant. With *Valence* and *Control/Power*, the learner can infer (un)pleasantness and the extent by which the colleague thinks she can act upon the situation. Finally, with *Valence*, *Control/Power* and, say, *Agency*, the learner can infer all of the above, but also obtaining cues about whether the causes of $x$ are to be looked into the actions of the colleague, or if the learner herself is somehow responsible.

Once again, the decision of striving for immediacy or accuracy in self-reporting emotion must be assessed in terms of pedagogical implications, rather than available user interfaces. A computational model that allows to make this decision by minimizing the impact on the technical requirements is therefore also pertinent to an *emotional awareness*'s interest. The next section thus describe the logic of model.

## The Foundations of the Computational Model Based on a *k*-Nearest Neighbors Approach {#foundations-computational-model}

This section first proposes a brief and general outline of computational models of emotion, used to situate the model at hand. Then, the model is described in terms of input, treatment, and output. Finally, the section lays out the requirements for the model to be adopted. Both description and requirements are indicated at a conceptual, abstract level. The concrete implementation in a programming language is available in the Code and Data Section at the end of the Introduction.

### Computational Models of Emotion

Computation models of emotion are a wide domain of inter-disciplinary research, spanning fundamental emotion theories and highly applied applications with computational agents endowed with emotions [@marsellaComputationalModelsEmotion2010; @mortillaroEmotionsMethodsAssessment2015; @calvoAffectDetectionInterdisciplinary2010; @guestHowComputationalModeling2021], see also Section \@ref(affect-aware-systems) about affect-aware systems. Compared to emotion theories more generally, which may sketch some inner functioning of emotion in broad terms while waiting for further evidence, computational models must instantiate the mechanisms in concrete terms, which are interpreted or compiled by a machine (*ibid.*). A positive turn of this necessity is that computational model can be presented -- and thus scrutinized -- in the smallest details.

@marsellaComputationalModelsEmotion2010 identify three main areas to which computational models of emotion may contribute. First, they can have an impact on fundamental theories of emotion, for instance from the psychological perspective. Second, they can empower the field of artificial intelligence and robotics. And third, they can contribute to human-computer interaction both at the intra-personal and inter-personal levels. The computational model set forth in this chapter is particularly driven from this last perspective, but may also contribute -- to a lesser extent though -- to the first. Furthermore, @marsellaComputationalModelsEmotion2010 also suggest that computational models, especially from a componential driven perspective, should make clear which components are included in the model. As emerged in the previous chapters of the thesis, the computational model adopted by the DEW is focused on the appraisal and categorization/integration modules.

Finally, congruently with @cerneaSurveyTechnologiesRise2015 list of emotion-related technologies, the computational model is meant to be parsimonious by requiring minimal hardware and also limited computational power.

### Model's Description

The aim of the model is to first sort, and later subset, a number of lexicalized emotions, which are provided as representative of the *true* emotional episode. To obtain this result, the computational model is based upon the following parameters, which are presented hereafter in a notation adopted especially throughout the chapter, and occasionally in the reminder of the manuscript:

-   $E$ stands for *Appraisal **E**valuation*, that is, the hypothetical value that a cognitive evaluation can have on a given set of appraisal criteria {$E_1$, $E_2$, ..., $E_i$}. In concrete terms, $E$ represents the rating of a person asked to evaluate one or more appraisal dimensions. If there was only one appraisal, *Novelty* for instance, than $E$ assumes a value determined by the extent to which the person evaluates the event as sudden, familiar or unpredictable on $E_1$. In terms of the DEW interface, $E$ refers thus to the slider(s), which prompt respondents to actively rate the situation against an appraisal/dimensional criterion. There in only one $E$ at a given moment $t$, but $E$ can be repeated at $t_1, t_2, t_3, ...$, each time assuming either the previous value or a different value.
-   $P$ stands for *Subjective Feeling **P**redicted Appraisal*, that is, the predicted appraisal profile associated with a given subjective feeling, with subscripts {$P_1$, $P_2$, ..., $P_j$} representing a set of $j$ predicted appraisal criteria. The *coordinates* of $P$ are identified with a common label, the lexicalized/discrete emotion, which is meant to be representative of the subjective feeling. There may be any number of labelled $P^*$ belonging to an affective space $A$, that is $P^a, P^b, P^c, ... \in A$. In concrete terms, each $P$ represents the appraisal profile that is supposed to elicit the given subjective feeling. For instance, the CPM predicts that a subjective feeling represented by the word *happiness* may be related to a higher evaluation on *Novelty* appraisal groups compared to a subjective feeling represented by the word *sadness* [@schererComponentProcessModel2010]. In term of the DEW functioning, $P$ is determined by the underlying affective space $A$ used as a *reference frame*. Contrary to $E$ which may varies according to $t$, $P$ holds constant for any given $A$ as long as $A$ is adopted as the *reference frame*.
-   Finally, $\Delta(E, P)$ represents the distance between the Appraisal Evaluation and the Subjective Feeling Predicted Appraisal, with the *k-nearest neighbors* being the *k* lexicalized emotions with the lowest $\Delta(E, P)$. In other words, if $k=5$, then, among all the options in the affective space, the 5 whose appraisal profile $P$ minimize the distance from $E$ represent the subset of most likely subjective feelings to occur, given the *mysterious* point $E$ provided by the person's cognitive evaluation. $\Delta(E,P)$ is computed at any variation of $E$ at time $t$, but holds constant if $E_{t1} = E_{t2}$.

The computational model therefore consists in a function that takes the three parameters $E$, $A$, and $k$ as Input, and provides a subset of $A$, consisting in a set of ordered $\overrightarrow{P}$, where the order is determined by $\Delta(E,P)$, as Output. In a more succinct form:

$$
f(E, A, k) \Rightarrow \mid\overrightarrow{P}\mid = k : P \in A \: \textrm{and} \: \Delta(E,P) \: \textrm{of} \: P^x \le \Delta(E, P) \: \textrm{of} \: P^{x+1}\
$$

### Model's Requirements {#model-requirements}

The computational model has minimal specific requirements. First, it is mandatory that the cardinality of $E$ is inferior or equivalent to the cardinality of $P$. That is, if respondents are asked to evaluate the situation on three appraisals/dimensions {$E_1, E_2, E_3$}, then the underlying affective space must be three-dimensional or higher. It must in fact provide at least coordinates for {$P_1, P_2, P_3$}, with the additional dimensions $P_4$...$P_j$ not adopted as *reference frames*.

Second, it is necessary that the domain of $E$ -- the possible values of the rating -- matches the domain of $P$. If the evaluation is rated in a domain ranging from [$V_{min}$:$V_{max}$], the lexicalized emotions in the affective space must be disposed somewhere between $V_{min}$ and $V_{max}$. Any values in the range can be potentially acceptable by the model, including the very same values for all options. In which case, though, the subsetting algorithm would be useless, since every lexicalized emotion would be always at the very same distance from $E$. But as long as at least two options in the affective space do not have exactly the same $P$, than the computation model guarantees to sort the suggested subjective feelings according to the minimal value of $\Delta(E, P)$.

The following sections thus illustrate how this distance $\Delta(E, P)$ is calculated by the model depending on the number of appraisal criteria/dimensions composing the underlying affective space: one-dimensional, two-dimensional, or multi-dimensional. The presence of a specific section on two-dimensional affective spaces -- which are in fact the first multi-dimensional condition -- is due to the possibility of interpreting the affective space either as a circumplex or as a Cartesian plane, which require a slightly different computation. The specific interest of each *n*-dimensional space are briefly discussed with respect to the existing literature.

## One-Dimensional Affective Spaces {#one-dimensional-affective-spaces}

In a one-dimensional affective space, where there is only a single appraisal criterion, the nearest neighbors are determined by the absolute distance between $E_1$ and $P_1$:

$$
\Delta(E, P) = |E_1 - P_1|
$$

One-dimensional affective space are not usually defined as such in the literature, even though the possibility to conceptualize emotion on a single dimension is not infrequent, especially when this dimension represents an omnibus *Valence* [@shumanLevelsValence2013; @erbasRoleValenceFocus2015]. The possibility to use a one-dimensional affective space is also warranted by the working definition proposed by @mulliganWorkingDefinitionEmotion2012 -- *i.e.*, *at least one appraisal* -- as well as by researchers and practitioners' habit to recur to a dichotomous categorization [@harleyDevelopingEmotionAwareAdvanced2017; @pekrunPositiveEmotionsEducation2002]. Even though the actual rating may be implicit rather than explicit, for instance in grouping forced choices in a *positive* or *negative* columns as in @molinariEmotionFeedbackComputermediated2013, this is still a uni-dimensional space in disguise.

The possibility to restrain the affective space to a single dimension may also be useful in situations where researchers or practitioners are interested specifically in the assessment of only one criterion, avoiding thus the potential mediation or moderation effects of additional criteria. For instance, if one is interested in the *Normative significance*, more specifically with respect to *External standards* -- whether the event violated laws or socially accepted norms [@schererWhatAreEmotions2005] -- the presence of another appraisal criteria such as *Valence*, may influence that evaluation. For example, negatively valenced emotion may be considered less socially acceptable [@vankleefInterpersonalDynamicsEmotion2018]. The presence of an explicit reminder of the valence could therefore mediate or moderate the rating of the external standard of normative significance.

## Two-Dimensional Affective Spaces {#two-dimensional-affective-spaces}

Two-dimensional -- or bi-dimensional -- affective spaces are ubiquitous in research, especially in the form of the dimensional *Valence x Arousal* circumplex [@russellCircumplexModelAffect1980], but also in appraisal-driven perspectives, as in Pekrun's Control-Value theory of achievement emotions [@pekrunControlValueTheoryAchievement2006; @pekrunControlValueTheoryAchievement2014] or the V*alence x Control/Power* structure of the Geneva Emotion Wheel [@schererGRIDMeetsWheel2013; @schererWhatAreEmotions2005]. As illustrated in the first part of the thesis, several scholars have theoretically adopted two-dimensional affective spaces in their research, but also as graphical representations for self-report/awareness tools [*e.g.*, @feidakisEndowingElearningSystems2011]. Consequently, it is important that an overarching computational model can manage to incorporate different kinds of two-dimensional affective spaces.

In this regard, an affective space based on two dimensions can be interpreted as the sum of a pair of one-dimensional continuous spaces, in which case each subjective feeling is determined by a precise coordinate in a Cartesian plane. Alternatively, it may also be interpreted, less stringently, as a circumplex where each subjective feeling is positioned at an equal distance from the origin, as in the GEW. The alternative interpretations of a two-dimensional affective space, depicted in Figure \@ref(fig:two-dimensional-affective-spaces), require different computational models from a *k-nearest neighbors* perspective. This section, thus, first illustrates the computational model for a Cartesian plane, and then for a circumplex. Last, it compares the two computational models in a simulation using the same affective space depicted either as a Cartesian plane or a circumplex.

(ref:two-dimensional-affective-spaces-caption) Comparison between a two-dimensional affective space for which the *reference frame* is a Cartesian plane (left) or a circumplex (right). The X and Y values corresponds to the $E_1$ and $E_2$ values respectively.

```{r two-dimensional-affective-spaces}
#| fig.align="center", 
#| fig.cap="(ref:two-dimensional-affective-spaces-caption)",
#| fig.scap = "Affective spaces as Cartesian planes vs. circumplexes", 
#| out.width="48%", 
#| fig.show="hold"
knitr::include_graphics(
  c(
    here::here("./figure/dew/dew-vector-distance.png"),
    here::here("./figure/dew/dew-radial-distance.png")
  )
)
```

### The Affective Space as a Cartesian Plane {#the-affective-space-as-a-cartesian-plane}

When a second appraisal criteria is added, the resulting affective space can be conceptualized as a Cartesian plane, in which $P$ is represented by the coordinates $x = P_1$, and $y = P_2$. At the same time, the appraisal evaluation is also determined by coordinates $x = E_1$ and $y = E_2$. Consequently, the *k-nearest neighbors* may be determined using the Euclidean (or Pythagorean) distance, with the nearest lexicalized emotions having the smallest hypotenuse of a triangle with the length of the sides determined by the difference between $E_1$ and $P_1$, and $E_2$ and $P_2$:

$$
\Delta(E, P) = \sqrt{(E_1 - P_1)^2 + (E_2 - P_2)^2}
$$

### The Affective Space as a Circumplex {#the-affective-space-as-a-circumplex}

An alternative for two-dimensional affective spaces consists in the use of a circumplex. A circumplex allows researchers to dispose multiple lexicalized emotions on the edge of a circle, so that the similarity between two items is inversely proportional to their distance on the circumference. The circumplex is determined by an x-axis and a y-axis, usually delimited by opposite poles of the same constructs (*e.g.*, pleasant/unpleasant) or two constructs in antithesis (*e.g.* sadness/enthusiasm). The combination of the two axes creates four quadrants, usually crossing over the negative and positive poles of the two axes. The top-right quadrant is x-positive and y-positive; the bottom-right quadrant is x-positive and y-negative; the bottom-left quadrant is x-negative and y-negative; and the top-left quadrant is x-negative and y-positive.

Given its practical structure, a circumplex can be used to dispose discrete emotions determined by two appraisals criteria in a more *relaxed* way compared to a Cartesian plane. In fact, researchers can tentatively dispose items at an equal distance from each other, assuming the order of disposition on the circle as a sufficient connotation of the affective space. Since the computational model, though, does not use the circumplex as a graphical representation, but rather as a *database* to retrieve suggestions, the fact that the options are not equally spaced on the circumference is irrelevant from a usability standpoint. As a consequence, it is also possible to adopt circumplexes with overlapping lexicalized emotions, or where the distance between any two items on the circle is not necessarily the same, but rather a function of some theoretical or practical mechanism.

On the other hand, the use of a circumplex as the underlying affective space requires a slightly different approach in computing the distance between $E$ and $P$, because $P$ in a circumplex is best conveyed as an angle, so that $P_{angle} \in \mathbb{R} : 0 \le P_{angle} < 360$. A possible solution is to use $E_1$ and $E_2$ as points in a Cartesian plane, and compute the slope from the origin [0,0] of the plane. The slope can then be used to retrieve the angle from the x-axis, using the inverse tangent, or arctangent trigonometric function, and -- with some correction skipped here, see Fritz (2015) or the code repository for more details -- also obtain an $E_{angle} \in \mathbb{R} : 0 \le E_{angle} < 360$:

$$
E_{angle} = arctan(\frac{E_1 - 0}{E_2 - 0}) + correction = arctan(\frac{E_1}{E_2}) + correction
$$

Once retrieved the $E_{angle}$, the *k-nearest neighbors* can be determined using the same logic as in a uni-dimensional condition, by computing the absolute distance between $E_{angle}$ and $P_{angle}$:

$$
\Delta(E, P) = |E_{angle} - P_{angle}|
$$Conceptually, with this solution, it is worth considering the case when both $E_1$ and $E_2$ are at the neutral point 0, in which case $\Delta(E, P)$ is exactly the same for every subjective feeling, since any point on a circle is at the same distance from the origin^[The case is also computationally problematic, since it entails an impossible division by 0]. In other words, any subjective feeling is as likely to represent the neutral point as any other. When both $E_1$ and $E_2$ are 0, thus, the computational model attributes a virtual random value both to $E_1$ and $E_2$ only with the purpose to shuffle the set of subjective feelings and avoid thus that some subjective feelings are arbitrarily suggested more frequently with the neutral point.

Furthermore, the computed slope is a ratio between $E_1$ and $E_2$, which entails that some ratios are more frequent than others assuming a sufficiently large range between [$V_{min}$:$V_{max}$]. The more approximate nature of the circumplex disposition is therefore also reflected in a more approximate computation of $\Delta(E,P)$.

### Computational Comparison Between a Cartesian Plane and a Circumplex {#computational-simulation}

Taking advantage of the fact that the disposition of the 20 subjective feelings in the third version of the Geneva Emotion Wheel [@schererGRIDMeetsWheel2013] has been determined using the GRID instrument, the same affective space is available as a circumplex and as a Cartesian plane. As a consequence, it is possible to simulate the behavior of the computational model with the alternative bi-dimensional spaces. For the purpose of the simulation, the following parameters will be adopted:

-   The two appraisal dimensions (*Valence* and *Control/Power*) are implemented on a continuum ranging from -100 to 100, with a one-unit step;
-   The *k*-nearest neighbors is set to $k = 3$ and to $k = 10$ to assess how the two models behave when the number of suggested feelings varies;
-   The subjective feelings in the circumplex affective space are disposed so that each is equally distant from the previous and the following item, with an initial rotation for the first item of half the distance between each subjective feeling to avoid an option on the 0° -- see the right side of Figure \@ref(fig:pof-gew-radial-vs-vector);
-   The subjective feelings in the Cartesian plan are disposed according to the values retrieved from data available in @schererGRIDMeetsWheel2013, see the left side of Figure \@ref(fig:pof-gew-radial-vs-vector). The original value on a range from -1 to 1 has been multiplied by 100 to map the range of the simulation.

(ref:pof-gew-radial-vs-vector-caption) Disposition of the Geneva Emotion Wheel's 20 subjective feeling on a circumplex (left) and Cartesian plane (right) affective space. Reconstructed from @schererGRIDMeetsWheel2013.

```{r pof-gew-radial-vs-vector}
#| fig.align="center", 
#| fig.cap="(ref:pof-gew-radial-vs-vector-caption)",
#| fig.scap = "The Geneva Emotion Wheel as a Cartesian plane or a circumplex affective space, reconstructed from Scherer, Shuman, et al. (2013)", 
#| fig.align="center", 
#| out.width="100%"
knitr::include_graphics(
  here::here("figure/dew-rt/gew-radial-vs-vector.png")
)
```

The simulation is computed by looping the $E_1$ and $E_2$ ranges so that each possible combination between the two -- that is, [-100, -100], [-100, -99], ..., [0, 0], ..., [99, 100], [100, 100] -- generates a subset of $k = 3$ or $k = 10$ subjective feelings. Figures \@ref(fig:pof-simulation-3) and \@ref(fig:pof-simulation-10) show the results of the simulation. With a $k = 3$ subset, the radial distance of the circumplex shows a more homogeneous number of appearances of each subjective feeling in the subset compared to the vector distance of the Cartesian plane. The opposite happens with $k = 10$, where the vector distance results in more homogeneous number of appearances compared to the radial distance.

(ref:pof-simulation-3-caption) Results of the simulation comparing distance computed as a radial or vector distance with a subset of $k = 3$ subjective feelings.

```{r pof-simulation-3}
#| fig.align="center", 
#| fig.cap="(ref:pof-simulation-3-caption)",
#| fig.scap = "Simulation of the computational model with k = 3", 
#| fig.align="center", 
#| out.width="70%"
dew_rt.sim_gew_graph_3
```

(ref:pof-simulation-10-caption) Results of the simulation comparing distance computed as a radial or vector distance with a subset of $k = 10$ subjective feelings.

```{r pof-simulation-10}
#| fig.align="center", 
#| fig.cap="(ref:pof-simulation-10-caption)",
#| fig.scap = "Simulation of the computational model with k = 10", 
#| fig.align="center", 
#| out.width="70%"
dew_rt.sim_gew_graph_10
```

This simulation has only illustrative purpose, because the two affective spaces are only examples of an infinite number of spaces that can be conceived and therefore results may be very different with different configurations. Nevertheless, this kind of simulation may be interesting to carry out before adopting an affective space to assess whether there are subjective feelings that are more likely to be suggested by the system, because that could influence learners in choosing them.

## Three-Dimensional Affective Spaces and Beyond {#three-dimensional-affective-spaces-and-beyond}

As previously mentioned, there is growing empirical evidence in suggesting that one- and two-dimensional affective spaces are not enough to account for the variety of emotional experiences in semantic terms [@fontaineWorldEmotionsNot2007; @gilliozMappingEmotionTerms2016; @fontaineLinearNonlinearRelationships2021]. A four dimensional structure determined by *Valence*, *Power*, *Arousal* and *Novelty* seems better fitted for a dimensional representation of emotions (*ibid.*). It is worth mentioning in this regard, though, that @fontaineLinearNonlinearRelationships2021 found empirical evidence suggesting that the additional *representative* value brought by a third and a fourth dimensions depends on the specific discrete emotions. For instance, terms related to *joy* and *sadness* could be discriminated almost as well with two dimensions as with four. On the other hand, lexicalized emotions closer to *anger* or *surprise* benefited from a four-dimensional structure [@fontaineLinearNonlinearRelationships2021].

In appraisal theories more specifically, there is virtually no limits to the number of appraisal criteria [@moorsAppraisalTheoriesEmotion2013]. As a reminder from Section \@ref(cpm-appraisal-module), the Component Process Model postulates as many as 13 sequential evaluation checks. As a consequence, an overarching computational model should be able to integrate higher-order dimensional spaces, where the number of dimensions can be determined by researchers and practitioners with respect to theoretical, practical, or usability criteria (*e.g.*, if the moment-to-moment temporal dimension is particularly important, asking learners to repeatedly rate a high number of criteria would be very time consuming).

Extending the model to more than two dimensions is relatively straightforward when the Euclidean distance is maintained as the algorithm to determine the *k-nearest neighbors*. In fact, it is possible to determine the distance between any two points in a multidimensional space simply by adding additional dimensions to the Pythagorean formula:

$$
\Delta(E, P) = \sqrt{(E_1- P_1)^2 + (E_2 - P_2)^2+\cdots+(E_n - P_n)^2}
$$

The model therefore remains fairly limited in term of computational power required, even if the number of dimensions increases. The difficulty with a higher-order affective spaces is in fact more conceptual than practical, since it is more difficult to imagine a space with three or more dimensions. Beyond three dimensions, it is even difficult to represent the space graphically, which contrasts with the easy-to-understand, and easy-to-implement disposition of subjective feelings in a circumplex. At the same time, the *k-nearest neighbor* approach can precisely contribute to overcome this issue: once a multi-dimensional affective space is established, the dimensions are *flattened* on uni-dimensional ratings. The computational model is then in charge of mapping uni-dimensional ratings into multi-dimensional distances, as vectors between points in spaces the human mind can hardly imagine from a spatial point of view.

As already mentioned, Gillioz and colleagues [-@gilliozMappingEmotionTerms2016] adopted the GRID instrument -- in its shorter CoreGRID version [@schererCoreGRIDMiniGRIDDevelopment2013] -- to assess the mapping of 80 emotion terms in an affective space. The number of dimensions and of emotion terms composing the affective space empirically established by the authors is a representative example of how a parsimonious computational model could be deployed in a self-report instrument, allowing users to benefit of all the dimensions and all the suggested emotion terms.

## Limitations Of the Computational Model and Possible Ways to Attenuate Them {#limitations-of-the-computational-model-and-possible-ways-to-attenuate-them}

The computational model presented here is overtly simple to account for the sheer complexity of *what an emotion is*. It is limited to the appraisal and integration/categorization modules, excluding thus three out of five components of the Component Process Model. The complexity of the two modules is also simplified. In absolute terms, thus, the model reflects only a fraction of the theoretical work of the CPM, appraisal theories, or emotion theories more broadly. Compared to other existing solutions, though, the model adds some distinctive features. The limitations of the tool are therefore assessed with respect to these distinctive features.

A first limitation consists in the need for the model to dispose of an accurate *reference frame*. Compared to a pure dimensional approach, a pure discrete emotion approach, or a combined -- but not integrated -- approach, the model highly depends on the specific details of the underlying affective space. As a reminder, for instance, the EMORE-L [@molinariEMORELOutilReporting2016] combines dimensions and discrete emotions, but without a link between the two. The *emot-control* [@feidakisProvidingEmotionAwareness2014; @feidakis2013] and Mood Meter application [@brackettRULERTheoryDrivenSystemic2019; @hoffmannTeachingEmotionRegulation2020; @nathansonCreatingEmotionallyIntelligent2016], as well as the Geneva Emotion Wheel, combine the two but only from a dispositional point of view, with discrete emotion that are nicely *spread* over the space to an equal distance for graphical ease of use. With the computational model at hand, the precise coordinates of each discrete emotion on the underlying affective space is taken into account *as is*. The simulations in Section \@ref(computational-simulation) have already highlighted this limitation. When the lexicalized emotion do not *occupy* the whole range from $V_{min}$ to $V_{max}$, there are discrete emotions that will appear more often in the subset, especially the ones at the edge of isolated clusters. Those will still be considered *nearest neighbors* when $E$ is actually in uncharted (*i.e.*, blank) territory.

This constraints impacts the model in another way. In fact, the model subsumes a linear relationship between the dimensions and the corresponding subjective feeling. Using statistical jargon widely adopted in psychology, the model does not take into account the possibility of *interactions*, whereas some subjective feelings may in fact be more likely to occur given a dimension conditional to the value of the other [@fontaineLinearNonlinearRelationships2021]. For instance, a subjective feeling on a dimension may be more probable when the evaluation of another dimension is either very high or very low, but not *neutral*. On the contrary, the model expects the subjective feeling to be monotonically (*i.e.*, proportionally) linked to both dimensions. @fontaineLinearNonlinearRelationships2021 have tested this possibility on the four-dimensional structure already depicted. They indeed found some interactions, especially when the third and fourth dimension are stacked in.

Given the intrinsic simplicity of the computational model, it was inevitable that some limitations occur. Any computational model must be calibrated around a ratio between the quality/reliability of the information *in* and the quality/reliability of the information *out*. [@mcelreathStatisticalRethinkingBayesian2020]. The flexibility of the computational model, though, can also be exploited to attenuate the limitations. Existing affective spaces are mainly driven from the theoretical perspective, meaning that they strive for accuracy. Depending on researchers and practitioners' specific aims, one can contemplate the possibility to apply a function $f(P)$ that transforms the affective space, for instance by *distancing* the discrete emotions as much as possible. The $f$ applied may be linear, but also non-linear, which may be useful for instance if one wants to merge existing affective dimensions into composite criteria [@fontaineLinearNonlinearRelationships2021]. This possibilities are nevertheless not explored in the present contribution.

Other ways to attenuate the limitations consists in adapting the interface of the DEW. As seen in Section \@ref(dew-interface), the tool can also integrate an "Other..." field, through which respondents can provide their own subjective feeling. With this option, the computational model is by-passed by respondents own interpretation of which lexicalized emotion represents at best their own subjective feeling.

## Summary

This chapter illustrated a parsimonious computational model at the core of the Dynamic Emotion Wheel. The model takes an *n*-dimensional affective space as an underlying *reference frame*, and compares the actual evaluation made by users with the predicted value of the pre-defined discrete emotion in the affective space. The model then computes, using a *k*-nearest neighbors approach, the discrete emotion who minimize the distance between the actual evaluation and the predicted value. These become the *k* suggestions that the model propose to the user as the most likely subjective feeling corresponding to the given dimensional evaluation. The computational model is therefore a form of *helper* or affective assistant, which gives *informed suggestions*, but that also reckons its limits. As a consequence, it is all the more important that the interface of the tool may be determined as much as possible from researchers and practitioners themselves, which is the subject of the next chapter.
