# A Parsimonious Computational Model Linking Appraisal and Subjective Feeling

```{r computational-model-setup, include=FALSE, echo=FALSE}
library(tidyverse)
library(papaja)
library(here)
library(knitr)
library(kableExtra)

# Load the relevant data and graphs
source(here("data", "dew-rt", "dew-rt-export.R"))

```

This chapter illustrates a parsimonious computational model inspired by the Component Process Model (CPM) theoretical framework that links appraisals criteria and subjective feeling. As a reminder, appraisals criteria in appraisal theories of emotions -- and the CPM more specifically -- are cognitive evaluation of an event, which determine emotion elicitation and differentiation. The subjective feeling, on the other hand, is the result of a highly dynamic and cumulative process, instantiated by appraisal, that coalesces in an overall emotional experience, from which the person may extrapolate meaning by recurring to natural language words or idioms that best convey its ineffable nature. In previous works, I have set forth the foundations of a prototype of an EAT, named *Dynamic Emotion Wheel* for its relationship with the Geneva Emotion Wheel illustrated in Part I of the thesis, which already explored the possibility to use active ratings of appraisal criteria to determine a set of subjective feelings most likely to occur, given the specific evaluation of the appraisal criteria. Previous work was nevertheless mainly driven by usability concerns emerged in empirical contributions using a provisional EAT [@molinariEmotionFeedbackComputermediated2013]. As it is often the case with usability issues, though, they were only the tip of the iceberg of more fundamental issues to be addressed, which initialized the quest for a better concept of an EAT. This chapter therefore builds on previous work in an attempt to bring an EAT even closer to appraisal theories of emotion, and the CPM more specifically.

The chapter starts by briefly describing previous work on the Dynamic Emotion Wheel. Then, it extends on it by implementing a *k-nearest neighbor* approach to the relationship between appraisals and subjective feeling. Harnessing the abstract concept of an *N-dimensional* affective space, a parsimonious computational model can link one or more appraisal criteria with a set of subjective feelings that ar most likely to correspond with the specific appraisal profile. While presenting the abstract model, existing affective spaces will be used to illustrate the concept in more concrete terms. At the end of the chapter, the core principle of an appraisal-driven EAT will emerge, paving the way to the possibility of configuring it according to theoretical or practical needs.

## The Dynamic Emotion Wheel

During the Master of Science in Learning and Teaching Technologies at Geneva University between 2013 and 2015, I had the chance to collaborate to the *Emotion Awareness Tool for Computer-Mediated Interactions* (EATMINT) project, which at the time had just concluded empirical investigation on the use of Emotional Awareness Tools in computer-mediated collaboration -- which then led to several publications [*e.g.*, @cereghettiSharingEmotionsComputermediated2015; @Chanel2013; @molinariEmotionFeedbackComputermediated2013]. The project had two intertwined main objectives:

1. Designing Emotion Awareness Tools, which allow either the explicit or the autonomic sharing of
emotions among participants in a computer-mediated task;
2. Studying the impact of emotional awareness on collaboration both at intra- and inter-personal
levels, with respect to the emergence of strategies that improve collaboration such as monitoring,
regulating and reflecting on emotions.


## A Parsimonious Computational Model Based on a *K-Nearest Neighbor* Approach

### One-Dimensional Affective Spaces

### Bi-Dimensional Affective Spaces

```{r}
dew_rt.sim_gew_graph_3
```

```{r}
dew_rt.sim_gew_graph_10
```

### Multi-Dimensional Affective Spaces

## Synthesis



