---
bibliography: references.bib
---

# (PART) Concluding Remarks {.unnumbered}

# General Discussion {#general-discussion}

```{r setup-general-discussion}
library(dagitty)
```

The general discussion of the thesis draws both from the implementation of the proof of concept in Part II and the results of the empirical contribution in Part III. It will be organized on what I consider to be the main contributions of the present work: the computational model at the core of the proposed Emotion Awareness Tool and the abstract model of emotional awareness presented in Section ... The first part of the discussion is therefore directed towards the role the computational model has played and may play in shaping how emotional awareness can be fruitfully conveyed from an appraisal-driven perspective. The second part of the discussion attempts to extend the abstract model of emotional awareness to a more concrete structural causal model. The discussion then proposes a series of suggestions that may be of interest to other scholars interested in implementing emotion awareness tools, which are derived from the overall experience in planning and developing the tool as well as situations in which to deploy it. Finally, the discussion ends with openings on future perspectives of the current work.

## Is Emotional Awareness a More Defined Construct?

The first objective of the thesis was related to the construct validity of emotional awareness as a concept. The construct as so far been defined mostly implicitly following what I think is a mild form of *psychologisme*: given that emotion play a prominent role in everyday life and it is also related to mechanisms that are implicated to learning, then it should also play a prominent role in computer-mediated learning environments. I reckon that in a recent and forming field of research, some amount of implicit assumptions is not only necessary, but also instrumental in attempting to determine where research should be directed [@fiedlerCycleTheoryFormation2004].

In this regard, I posited that the whole concept of emotional awareness could be better grasped if learners (1) reflect on their emotion from a subjective feeling experience rather than autonomic recognition; (2) guide their reflection on the appraisal of a situation; and (3) convey emotional awareness to others using the same *information unity* they have used to extrapolate meaning-making from the situation at hand. Following these assumptions, I set forth two models that attempt to advance our understanding of emotional awareness in computer-mediated learning environments: an abstract model of emotional awareness, and a computational model that assists learners in bridging the cognitive evaluation of the situation with their subjective feeling. I propose thus a comprehensive analysis of the two models.

### Analysis of the Abstract Model

The abstract model, illustrated again in Figure \@ref(fig:thesis-scm-model-repeated), consists in reducing emotional awareness to a core of simpler functions, which are put into a common perspective. It starts from the learning activity as a source of potential emotional experiences and paves two parallel paths at the intra-personal and inter-personal levels, culminating in meaning-making, which can be related back to the learning activity.

(ref:thesis-scm-model-repeated-caption) Abstract model of the functions of an EAT from the perspective of a single learner disposing of it in a computer-mediated learning environment. Repeated from Figure @ref(fig:thesis-scm-model).

```{r thesis-scm-model-repeated}
#| fig.cap="(ref:thesis-scm-model-repeated-caption)", 
#| fig.align= "center",
#| fig.scap = "Abstract model of emotional awareness", 
#| out.width= "100%"
knitr::include_graphics(
    here::here("figure/intro/thesis-scm-model.png")
)
```

The model has, to my sense, the merit of uncovering some of the implicit assumptions that are made in considering emotional awareness instrumental to computer-mediated learning environments. In particular, it draws a clearer distinction between the intra-personal and inter-personal perspectives, going thus in the direction of "disentangling the effect of reflecting upon one's own emotions from the effect of sharing one's emotions with the partner" [@molinariEmotionFeedbackComputermediated2013, p.342]. Highlighting more clearly the intra- and inter-personal building blocks of emotional awareness paves the way to the possibility of configuring tools and environments in which the different forms of emotional awareness are more or less explicitly sought after and enhanced.

The experiment in which the use of and access to emotional information was manipulated is a first attempt in this direction. Even if the results have been undermined by the small sample size at hand, the experiment at least succeeded in pointing out that, in the definition of emotional awareness as "information about one's own emotions and/or about the emotions of other learners", the *and/or* disjunction has profound consequences on the mechanisms at play in producing and perusing emotional awareness. The three word clouds used in the longitudinal study to convey persistence of the emotional experience at the personal, inter-personal, and group levels are another example of the disjunction. They directly mobilize the affiliation and distancing function of emotion identified by Fischer and Manstead [-@fischerSocialFunctionsEmotion2016]. Finally, the factor analysis on the Emotion Awareness Usefulness (EAU) scale, also introduced in the longitudinal study, corroborates the presence of a self-centered cluster (*Self-Understanding* and *Self-Regulation dimensions*) and a partner(s) oriented cluster (*Social Presence*, *Understanding Others*, and *Self-Other Comparison* dimensions). In other words, the distinction between the intra- and inter-personal paths is not only pertinent in the concrete, technical functioning of the EAT, but also in the perceived usefulness. In this regard, empirical data collected in both studies seem to privilege an inter-personal stance of the EAT, but with a *pure* intra-personal perspective which could also play a prominent role.

In the meantime, the abstract model falls short in purporting the left-hand side of the model: how meaning-making is extrapolated and related back to the learning activity. Even though it was not an explicit objective of the thesis to explore this relationship above and beyond the role of an EAT, this mechanism is at the core of why emotional awareness should be provided in the first place. The thesis made the overt assumption that, for this passage even to occur, it is important to maximize the chance that meaning-making is made on solid ground from an emotional perspective. I reckon that this assumption is open to valid criticism: whether the ground is solid should be determined, *in primis,* from a pedagogical perspective. A mainstream position on this subject is that learners can be guided by instructional designs or learning environments towards emotional experiences that are inherently instrumental to learning. I have chosen to take an alternative path by leveraging on the fact that emotions are in and of themselves adaptive responses to an environment that is likely to elicit some prototypical cognitive evaluations. Linking emotional awareness to appraisal theories of emotion is, to my sense, the best chance of bestowing learners with the responsibility of extrapolating useful meaning-making for their learning processes and outcomes from personal and inter-personal emotional information.

The overall use of the EAT made by learners in the empirical contributions of the thesis is, I think, reasonable support for the core of appraisal theories of emotion: even during similar -- if not identical -- situations, learners showed considerable variance in their evaluation and differentiation of their emotional experience. As a consequence, I consider that even in an instructional design that is tailored for inducing inherent functional and avoiding dysfunctional emotional experiences, learners could vary at different extent in *complying* with the paved emotional path. As a consequence, I join other scholars in considering that learners should be assisted in making the best out of emotion as a mechanism that provides important information on what is going on, paving the way to different opportunities in self-regulation, co-regulation and socially-shared regulation of learning [@jarvelaEnhancingSociallyShared2015; @jarvelaSociallySharedRegulation2016; @millerScriptingAwarenessTools2015; @winneWhatStateArt2015].

The sheer presence of an EAT seems to me, thus, a necessary condition -- though not sufficient, as the lack of use in the longitudinal study has pointed out -- for the whole abstract model to be put in motion. As pointed out by prior experiments and scholars, emotions are still a neglected mechanism in learning situations, and the presence of an EAT play therefore an important role in enhancing emotion alertness. On the other hand, the specific features of the EAT contribute to emotion conceptualization and, by extension, on the meaning-making activity that can be retrieved from the use of the EAT. To this extent, the explicit link between the cognitive evaluation and the resulting subjective experience is not only a technical feature that reduces complexity on screen. It is a central, pedagogical feature that is meant to facilitate the derivation of patterns "if I evaluate the situation as such as such, then I am more likely to feel such and such emotional experience" or, conversely, "if my partners are feeling in such or such a way, they are more likely to have appraised the situation in such or such way". The analysis of the computational model presented hereafter must therefore take into account the whole responsibility of the model.

### Analysis of the Computational Model

The computational model at the core of the proposed EAT is the mechanism by which this broad assumption is put in concrete terms: learners evaluate the situation on a set of appraisal criteria and are assisted in finding the subjective feeling that best convey their emotional experience based on theoretical ground. The evidence gathered in the empirical contributions of the thesis sketches a contrasting picture with respect to this assumption, as highlighted in the data-driven assessment of Chapter \@ref(study-3). On the one hand, the computational model based on the Valence x Control/Power appraisals obtained a fair predictive accuracy as an expert system. Most of the subjective feeling expressed belonged to the subset of computed suggestions. On the other hand, the use of the tool confirmed previous work in highlighting a major overlapping between the two dimensions, which tend to co-vary. As previous scholars have pointed out, it is safe to assume that this is mainly due to the overarching importance given to the Valence dimension, with the Control/Power following suit, conditioned on whether the situation is perceived as pleasant or unpleasant.

This is clearly a major shortcoming both from a user experience and pedagogical perspective. For the former, the learner is asked for an explicit action that seems to add little independent value, which makes it redundant and in contrast with the general outlook in user experience advocating parsimony and relevance in prompting user action. For the latter, learners seem to gain little understanding from the Control/Value reflection, even though, at a theoretical level, this is a major determinant in discriminating emotional experience, and also a relevant pedagogical dimension in influencing learning processes and outcome.

The extension of the computational model to multi-dimensional affective spaces illustrated in Chapter \@ref(computational-model)was mainly driven by this major problem. Allowing researchers and practitioners to adopt the number and kind of appraisal dimensions, which also determine the likelihood of the subjective feeling, is certainly a way to overcome the issue. Depending on the specifics of the computer-mediated learning environment, there may be cognitive evaluations of the situation at hand that are more or less relevant. This leaves the door open to future work targeting more specifically emotions linked to learning, as it is the case, for instance, to epistemic emotion which have only recently been put under the radar.

At the same time, as already mentioned, the way the interface of the EAT is built makes the underlying computational model independent from the appearance of the tool. This paves the way to alternative computational models that can be implemented on the very same user interface. I identify three possible kind of computational models: theory-driven, data-driven, and pedagogy-driven. These are illustrated by adopting the same conventions adopted in illustrating the computational model in Chapter \@ref(computational-model) : $E$ represents the expected subjective feeling, and $A$ represents the set of appraisal criteria against which the situation is assessed.

The theory-driven approach is the one adopted in the current contribution, and is represented by the following equation:

$$
E = A + Theory
$$

In this approach, the likelihood of an expected subjective feeling is derived only by the theoretical model, represented by the affective space. In other words, the model is not influenced neither by contextual factors, nor by the information coming in from the use of the tool through time -- for instance, the previous emotion experienced by the learner. This way of considering an emotion is consistent with the abstract idea, particularly relevant in appraisal-driven theories, of an emotion as a *Swiss-knife* that maximizes its usefulness in many different circumstances, depending on the person making the appropriate evaluation of which *tool* is more appropriate to the situation at hand on a moment-by-moment basis.

A first alternative to pure theory-driven approach consists in harnessing the concrete use of the tool through time, with the possibility to add the information about the previous emotion experienced. The equation would therefore be the following:

$$
E = A + {Use} + {Previous}
$$

In this case, the computational model would not be an *expert system* with a fixed set of responses, but would adapt depending on the use that is made of the tool. With a platform built around the EAT, it would be possible to set up some form of machine learning that uses existing data as training set. The previous experienced emotion could also be adopted as a predictor, leveraging on the chances that there are transitions between emotions that are more likely than others to occur in learning contexts [@dmelloDynamicsAffectiveStates2012; @dmelloConfusionCanBe2014; @lehmanConfusionComplexLearning2012]. This approach does not exclude the possibility to inject theoretical knowledge, for instance as a form of Bayesian prior that regularizes the model [@mcelreathStatisticalRethinkingBayesian2020].

-   Pedagogy-Driven: the computation is based upon an expert system that calibrates the probability of having a defined evaluation and emotion experience given the specific of the task at hand. This is a sort of Bayesian interpretation: rather than having the cognitive evaluation free to vary as it is only determined by the person itself. In the theory driven we have $P(emotion|evaluation)$ whereas from an inverse perspective we have $P(evaluation|target \space emotion)$.\
    $E = A + {Scripting}$

Role of computational model overestimated in UX and underestimated in pedagogical terms.

## Is the EAT at Hand Instrumental in Research About Emotional Awareness in Computer-Mediated Learning Environments?

## Suggestions for the Implementation of Emotion Awareness Tools

1.  The tool should not only be anchored to an emotion theory, but the link should also be illustrated in terms of what are the advantages for learners to implement that theory rather than another.

2.  The tool should make explicit which kind of affective phenomena it targets

3.  The tool should justify the kind and number of dimensional/discrete emotion unit among which learners can choose

4.  

## Future Perspectives

### From an Abstract to a Structural Causal Model of Emotional Awareness

-   Brief introduction to SCM and DAG

-   Exemple of controlling for gender in CSCL experiments

-   Attempt to draw a preliminary SCM

### Technical Improvements to the EAT

### Further Empirical Adoption of the EAT
