---
bibliography: references.bib
---

# (PART) Concluding Remarks {.unnumbered}

# General Discussion {#general-discussion}

```{r setup-general-discussion}
library(dagitty)
library(rethinking)
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
```

The general discussion of the thesis draws both from the implementation of the proof of concept in Part II and the results of the empirical contribution in Part III. Since the previous chapters have already proposed extensive analyses of the specific topics or research questions at hand, I will adopt a higher-level perspective for the general discussion, focusing mainly on what I think are the two main contributions of the thesis: the abstract model of emotional awareness, and the computational model at the core of the proposed EAT. I then use the overall experience of implementing and assessing an EAT to provide what I think are useful suggestions for other scholars interested in providing emotional awareness through a dedicated tool.

## Analysis of the Abstract Model of Emotional Awareness

One of the main objectives of the thesis was related to the construct validity of emotional awareness as a concept. To my mind, the construct has so far been defined mostly implicitly following a mild form of *psychologisme*: given that emotion play a prominent role in everyday life and it is also related to mechanisms that are implicated to learning, then it is safe to deduce that it should also play a prominent role in computer-mediated learning environments. In a recent and forming field of research, some amount of implicit assumptions is not only necessary, but also instrumental in attempting to determine where research should be directed [@fiedlerCycleTheoryFormation2004]. In this regard, in an attempt to further clarify the concept, I have proposed an abstract model based on three underlying assumptions of emotional awareness: first, that emotion play a prominent role at the inter-personal level; second, that emotion convey useful information at the inter-personal level; and third, that emotion can be fruitfully expressed and perused in computer-mediated learning environments.

(ref:thesis-scm-model-repeated-caption) Abstract model of the functions of an EAT from the perspective of a single learner disposing of it in a computer-mediated learning environment. Repeated from Chapter \@ref(eat-general-chapter).

```{r thesis-scm-model-repeated}
#| fig.cap="(ref:thesis-scm-model-repeated-caption)", 
#| fig.align= "center",
#| fig.scap = "Abstract model of emotional awareness", 
#| out.width= "100%"
knitr::include_graphics(
    here::here("figure/intro/thesis-scm-model.png")
)
```

The abstract model, illustrated again in Figure \@ref(fig:thesis-scm-model-repeated), consists in reducing emotional awareness to a core of simpler functions, for which some potential intervening factors are provided. It starts from the learning activity as a source of potential emotional experiences and paves two parallel paths at the intra-personal and inter-personal levels, culminating in meaning-making, which can be related back to the learning activity. In this section, I propose a critical evaluation of the model by integrating support -- or lack thereof -- of empirical data.

### Disentangling Intra- and Inter-Personal Emotional Awareness

The model has, to my sense, the merit of uncovering some of the implicit assumptions that are made in considering emotional awareness instrumental to computer-mediated learning environments. In particular, it draws a clearer distinction between the intra-personal and inter-personal perspectives, contributing thus in "disentangling the effect of reflecting upon one's own emotions from the effect of sharing one's emotions with the partner" [@molinariEmotionFeedbackComputermediated2013, p.342]. Highlighting more clearly the intra- and inter-personal building blocks of emotional awareness paves the way to the possibility of configuring tools and environments in which the different forms of emotional awareness are more or less explicitly sought after and enhanced.

The experiment of Chapter \@ref(strudy-1), in which the use of and access to emotional information was manipulated, is a first attempt in this direction. Even if the results have been undermined by the small sample size at hand, the experiment at least contributed to point out how, in the definition of emotional awareness as "information about one's own emotions and/or about the emotions of other learners", the *and/or* distinction has profound consequences on the mechanisms at play in producing and perusing emotional awareness. Even though the overall experiment suggests that, as one may expect, inter-personal awareness primes over intra-personal awareness, there has nevertheless been also some indications of a potential use for individual awareness. In this regard, I consider very important to have some performance-based metrics, which allow to quantify the actual use of the tool. It is known that, in high-demanding tasks in particular, users of a technological artifact can have an altered perception of what they have been really doing with it. Especially with a dedicated tool that falls outside the content space of a learning task, performance-based indications are particularly revealing. To this extent, the measurement of eye-tracking transitions between zones of the computer-mediated learning environment seem to me a very promising indicator of the integration between emotional information and the learning process. In fact, it does not only provide a metric of the segregated peruse of each part of the screen, but also a dynamic overview of the potential links between information that users may be doing during cognitive processing.

In the longitudinal study of Chapter \@ref(study-2), the distinction between the two paths was less clear-cut being assured only by the three word clouds used to represent the affiliation and distancing function of emotion identified by Fischer and Manstead [-@fischerSocialFunctionsEmotion2016]. Nevertheless, some empirical data also allowed to reveal the two distinct paths. For instance, the evaluation of the Emotion Awareness Usefulness (EAU) pointed out how learners had higher expectations on the inter-personal dimensions compared to the intra-personal dimensions. The distinction on the scale was further corroborated by the presence, in the exploratory factor analysis, of a self-centered cluster (*Self-Understanding* and *Self-Regulation dimensions*) and a partner(s) oriented cluster (*Social Presence*, *Understanding Others*, and *Self-Other Comparison* dimensions). In other words, the distinction between the intra- and inter-personal paths is not only pertinent in the concrete, technical functioning of the EAT, but also in the perceived usefulness.

In this regard, the lack of use of the tool in the longitudinal study may derive from learners difficulty to match the two paths one against the other. The inter-personal path was blurred both graphically (with a grouped representation) and chronologically (without a common event that may be identified as source of the emotional experience). As a result, learners did not dispose of a *mirroring* image of other learners for the mutual modeling to occur [@dillenbourgSymmetryPartnerModelling2016]. This would be consistent with the assumption that others' emotions have value especially when they can be integrated as situated information for one's own behavior [@vankleefInterpersonalDynamicsEmotion2018; @fischerSocialFunctionsEmotion2016]. If the *pipe* of information stemming through the inter-personal channel cannot be incorporated in the intra-personal meaning-making, it may be difficult or impossible to project oneself truly in a shared socio-affective milieu [@jezegouCreerPresenceDistance2010; @kreijnsSocialAspectsCSCL2013; @rimeEmotionElicitsSocial2009].

I think it will be useful to insist in decomposing the paths in different situations and conditions, trying different versions of an EAT with respect to the emotional information it provides. In particular, the thesis hasn't tested a *pure* intra-personal setting, where emotional awareness is prompted even though there is no contact whatsoever with other learners. In the *Self-Centered* condition of Study 1, there was still some form of communication and social presence with another learner, even though a simulated one. Moreover, the participant knew -- or suspected at least -- that the other person was also reflecting on her emotions, since this was an overt request in the empirical task. In Study 2, even if the setting was individual on the learning task, the aim was for emotional awareness to contribute to an overall sense of *socio-affective* presence and sense of belonging. As long as there is some inter-personal contact in a self-report setting, the two levels are necessarily intertwined, since inter-personal awareness is conditioned to the production of intra-personal awareness. Consequently, it is difficult to rule out a form of *altruistic effect*: by thinking that one is mainly producing emotional awareness for others to peruse, one is actually benefiting from producing the data in the first place.

It would therefore be interesting to replicate the use of an EAT similar to that of the study by Molinari and colleagues [-@molinariEMORELOutilReporting2016] with the EMORE-L tool, depicted in Chapter \@ref(eat-general-chapter), but using the proposed EAT with the possibility to express emotions on a moment-by-moment basis. In Molinari and colleagues' study (*ibid.*), the authors highlighted how students praised for the possibility to dispose of inter-personal emotional awareness. In a direct comparison between the use of a similar EAT in a *pure* individual setting and a more socially oriented condition, it would be possible to quantify the different use and perception of usefulness of intra- vs. inter-personal emotional awareness.

### Passing From an Abstract to a (Difficult) Structural Causal Model of Emotional Awareness

The abstract model provides a framework through which emotional awareness is laid out in an organized manner. In programming, such a model would be qualified as an interface, which scaffolds the overall flow of information, but still requires concrete functions to be implemented for *something* to really happen. At the same level, the abstract model does not provide any concrete function that maps causes and consequences of emotional awareness. Alongside the abstract model, thus, it would be helpful to hypothesize structural causal models derived by the scientific knowledge about causes and consequences of emotional awareness.

A Structural Causal Model (SCM) combines a set of mathematical derivations about causal inference with graphical models, called Direct Acyclic Graphs (DAG), which allow to represent and infer causal relationships or independence between any two variables depending on how all the variables are connected in the overall model [@pearlIntroductionCausalInference2010; @pearl2016; @pearl2000; @pearl2018book; @rohrerThinkingClearlyCorrelations2018; @mcelreathStatisticalRethinkingBayesian2020]. A general example of an SCM as a DAG is proposed in Figure \@ref(fig:example-dag). Contrary to the abstract model presented in the thesis, were arrows mainly represent heuristics about some potential flow of information, in a structural causal model an arrow implies a functional relationship between the variable from which the arrow starts and the variable on the other end of the arrow. At the same time, the absence of an arrow implies independence between the two variables.

```{r example-dag}
#| fig.cap="General example of Direct Acyclic Graph (DAG)",
#| out.width="60%",
#| fig.align="center"
example_dag <- dagitty('dag {
bb="-4.44,-3.29,3.228,4.499"
A [adjusted,pos="-2,-1.5"]
B [pos="1.0,-1.5"]
C [pos="-0.5,1.5"]
F [pos="-0.5,-1.5"]
M [pos="-0.5,-0.5"]
X [exposure,pos="-2,0.5"]
Y [outcome,pos="1.0,0.5"]
A -> X
B -> Y
F -> X
F -> Y
M -> Y
X -> C
X -> M
X -> Y
Y -> C
}')

drawdag(example_dag, lwd = 2, goodarrow = FALSE)

```

One of the main advantages of SCM is that it is possible to derive, according to the specific effect that we want to test in the model, what other variables should be controlled/conditioned for depending on the type of effect we are interested in [@rohrerThinkingClearlyCorrelations2018; @pearlCausalInferenceStatistics2016; @cinelliCrashCourseGood2020a]. For instance, in the general example, we may be interested in the direct effect of $X$ on $Y$. In this case, the structure of the DAG implies that we must control for variables \$A\$, $F$, and $M$, but not for the others. Namely, controlling for $C$ would represent a *bad control*, since it would open another path from $X$ to $Y$ and thus potentially confound the direct effect. The control of $B$ is optional, depending on the degree of accuracy in estimating the effect $X \rightarrow Y$. Conversely, if we were interested in the total effect of $X$ on $Y$, then we must control only for $A$ and $F$, but not for $M$ and the others, with $B$ once again as optional. In other words, what variables are controlled for depend on the structural causal model itself.

An example taken from other contributions investigating emotional awareness may contribute to make the argument more concrete. For instance, in four experimental contributions investigating the role of emotional awareness in computer-mediated computer environments using dyads -- Eligio and colleagues [-@eligioEmotionUnderstandingPerformance2012], Molinari and colleagues [-@molinariEmotionFeedbackComputermediated2013], and Avry and colleagues [-@avryAchievementAppraisalsEmotions2020; -@avryExplicitSharingEmotions2021] -- all authors controlled for the gender of participants, either by taking only women or by forcing same-gender dyads. This experimental control is explained in @eligioEmotionUnderstandingPerformance2012 for the need control for differences in emotion understanding between men and women, whereas in @molinariEmotionFeedbackComputermediated2013 with the need to control for the effect of social roles in mixed dyads. It is not specified, though, how these phenomena could intervene in the process.

For instance, the four contributions are experiments in which there is a randomized trial. The strength of randomization is to severe all variables that can influence the choice of the intervention, that is, all potential causal arrows incoming into the independent variable (i.e. *with* or *without* emotional awareness). So it would not be useful to control for same-gender dyads if authors were worried this may influence the independent variable. On the contrary, conditioning on the independent variable (i.e. as variable $A$ in the general example) rather risks to reduce the direct effect on the dependent variable, since stratifying for only a part of the overall population reduces the potential variation at play. In other words, the control risks to undermine the very effect of interest.

At the same time, if the gender of the dyads may have an influence on the use of the EAT itself, for instance, on how many emotions are expressed by the user, than we are dealing with a post-treatment variable. In fact, the dyads without emotional awareness cannot be influenced by it, and therefore we are stratifying for something that is directly dependent on the intervention itself, which is also generally known to be a bad control.

Finally, same-gender dyads may have been controlled, as it is often the case in randomized trial, to have a better estimate of the effect on the outcome variable (i.e. as variable $B$ in the general example). In this case, thus, same-gender dyads need to be causally linked with the dependent variable. That is, if authors are interested in some form of measure about the learning gain in disposing or not of emotional awareness, same-gender dyads should influence the learning process or outcome itself. The four contributions used very different outcome variables in their experimental plans, so it is hard to believe that this control has been made for this reason.

I rather think that control for same-gender dyads has been added as a consequence of the unfortunate pressure researchers have or receive to *control for something,* for their study to seem more accurate. The practice nevertheless seems to have caught on and could easily be established as a must in the field, even though it is not explained causally. To be clear, I am not implying that the composition of the dyad with respect to the gender of learners (or other characteristics for that matter) is irrelevant. I am rather positing that the presence of controlling variables should be causally defined in terms of (a) an underlying structural causal model, and (b) the very effect the contribution is meant to explore. According to the interplay of these two elements, one can establish the necessary variable to adjust for.

This task is inevitably complex, considering the number of potential intervening factors at play, some of which have already been listed in the presentation of the model in Part I. We certainly need to make some trial and error before even getting close to identify a chain of causal variables due to the complexity of the phenomenon at hand. In the next section, I will therefore attempt to propose a tentative SCM based on the underlying logic and cues provided by the present contribution.

### A Structural Causal Model of the Effect of Emotional Awareness on Learning Processes and Outcomes

As mentioned above, a Structural Causal Model (SCM) depends on the specific effect we are interested in. For this purpose, I will adopt the overall question concerning the direct effect of emotional awareness on learning processes and outcomes. According to this aim, I have drawn the Direct Acyclic Graph (DAG) depicted in Figure \@ref(fig:tentative-dag).

```{r tentative-dag}
#| fig.cap="Tentative Structural Causal Model of the Direct Effect of Emotional Awareness on Learning Processes and Outcomes",
#| out.width="80%",
#| fig.align="center"
tentative_dag <- dagitty('dag {
bb="-3.883,-3.983,4.105,2.663"
"EAT Features" [pos="2.900,-3.2"]
"EAT Use" [pos="-1.379,-0.537"]
"Emotional Awareness" [exposure,pos="-2.5,0.5"]
"Emotional Competence" [pos="-2.5,-3.2"]
"Learning" [outcome,pos="3.027,0.5"]
Appraisal [pos="-2.5,-1.303"]
Partners [pos="0.973,-3.2"]
Regulation [pos="0.132,0.5"]
Task [pos="-0.929,-3.2"]
"EAT Features" -> "EAT Use"
"EAT Features" -> Appraisal
"EAT Use" -> "Emotional Awareness"
"Emotional Awareness" -> Regulation
"Emotional Competence" -> "EAT Use"
"Emotional Competence" -> Appraisal
"Emotional Competence" -> Regulation
Appraisal -> "EAT Use"
Appraisal -> "Emotional Awareness"
Partners -> "EAT Features"
Partners -> "EAT Use"
Partners -> Appraisal
Partners -> Regulation
Partners -> Task
Regulation -> "Learning"
Task -> "EAT Use"
Task -> "Learning"
Task -> Appraisal
Task -> Regulation
}')

drawdag(tentative_dag, lwd = 2, goodarrow = FALSE)
```

The SCM makes the following assumptions:

-   There are four high-level variables, at the top of the model, which play a prominent role: (1) the EAT features, (2) learners' Emotional Competence, (3) the characteristics of the Task, and (4) the presence/characteristics of the Partner(s). Among the four high-level variables, I posit that there is an influence of the Partner(s) on the Task at hand and the EAT Features, since the presence of Partners determines what learners will do, but also how the EAT would look like (e.g. configuration according to intra-personal, inter-personal, or both form of awareness). On the contrary, there is no link between the Task and EAT features, since the proposed EAT is meant to be multi-purpose, and therefore adapt to any task.

-   Congruently with the appraisal-driven perspective of the thesis, I assume that the Appraisal variable play a central role in the model. Appraisal is influenced by all the four high-level variables depicted above. In fact, how the situation is appraised is determined by the learner's Emotional Competence, the specifics of the Task at hand, the Partner(s) -- according to the Emotion As Social Information framework [@vankleefInterpersonalDynamicsEmotion2018] -- and the EAT features, which can guide the appraisal for instance through the chosen criteria. Appraisal is then a direct ascendant of Emotional Awareness, since I venture that a great deal of variation in Emotional Awareness is determined on how the situation is evaluated. The effect of Appraisal on Emotional Awareness can be direct or mediated by EAT Use. In fact, it is possible to construe Emotional Awareness even without using the EAT, for instance by inferring emotional meaning from cues outside the emotional information provided by the EAT [@buderGroupAwareness2021; @janssenCoordinatedComputerSupportedCollaborative2013].

-   The EAT Use also plays a prominent role in the model, since it directly influences Emotional Awareness. At the same time, the use of the EAT is influenced by the four high-level variables in a way that is very similar on how the four high-level variables determine Appraisal. The distinction between Appraisal and EAT Use is therefore determined mainly by the fact that (a) learners can decide not to use the EAT in spite of a relevant Appraisal, and (b) learners can convey or infer (deliberately or for lack of competence) an emotional information that is not congruent with the Appraisal [@molinariEMORELOutilReporting2016; @lavoueEmotionAwarenessTools2020; @scherer2021].

-   Emotional Awareness is determined both by Appraisal and the EAT Use, for reasons specified above. At the same time, Emotional Awareness directly determines the (emotion) Regulation strategy adopted, for instance Self-Regulation, Co-Regulation or Socially-Shared Regulation [@millerScriptingAwarenessTools2015; @grossEmotionRegulationCurrent2015; @reeckSocialRegulationEmotion2016; @jarvelaSociallySharedRegulation2016]. The Regulation also depends on three out of four high-level variables, since I venture that the EAT Features do not have a direct influence on Regulation. The influence of the EAT on Regulation is completely mediated by the path going through EAT Use and Emotional Awareness.

-   Finally, Learning Processes or Outcomes are directly determined by the Regulation, as well as by the Task, since, for instance, if the Task is too difficult, there will not be any Learning regardless of all other variables implied. On the contrary, I speculate that the influence of the Partners on Learning is all mediated by the other paths, since a form of interaction is required [@dillenbourgSymmetryPartnerModelling2016].

The structure of the DAG implies that, in spite of the multiple associations between variables, for measuring the direct effect of Emotional Awareness on Learning Process and Outcomes, it is necessary to control only for the Task and Regulation variables. The Task is fairly obvious, since to measure the effect of different degrees of Emotional Awareness the Task should be hold constant. The Regulation, on the other hand, cannot be hold constant. So an alternative could be to measure learner's regulatory abilities, for instance with the regulation and management dimensions of the Geneva Emotion Competence Test [@schlegelGenevaEmotionalCompetence2018] adopted in the longitudinal study.

It is worth pointing out, though, that the proposed SCM considers learning processes and outcomes in a very abstract way, a point that will be more thoroughly assessed in the limitations of the overall thesis in the Conclusion. As a consequence, other variables may need to be added depending on the specific nature of the outcome variable.

For instance, if the outcome variable is mainly related to the *content space* of the learning environment, such as information retention or problem-solving performance [@avrySharingEmotionsContributes2020; @avryExplicitSharingEmotions2021], added variables should probably adjust for antecedents of the dependent variable in an attempt to increase precision on measuring the effect. The proposed model and adjustements may thus stand as it is for the left-hand side of the structure.

On the other hand, if the outcome variable is more related to the *relational space* of the learning environment, such as quality of interaction [@molinariEmotionFeedbackComputermediated2013], other socio-affective phenomena could also need to be adjusted, especially if they influence both the outcome variable and Regulation. If one draws an arrow from the Partner to the Learning Processes and Outcomes and maintains the arrow from the Partner to Regulation, for example, then it is necessary to control for Partner if the direct effect is the aim of the study (c.f. the discussion about same-gender dyad in the previous subsection).

### Synthesis

Emotional Awareness in Computer-Mediated Learning Environments is linked to many potential intervening factors. As a consequence, it may be useful to reveal the assumptions of causality between those variables in a concrete manner. Starting from the abstract model proposed in Chapter \@ref(eat-general-chapter), useful as a general interface for organizing emotional awareness, I have derived a structural causal model that makes explicit assumptions about the underlying structure of a direct effect of emotional awareness on learning processes and outcomes. In the structural model, the Appraisal and EAT Use variables are at the forefront of many interconnection of the model. In the proposed EAT, the two variables coalesce in the computational model, which is discussed next.

## Analysis of the Computational Model at the Core of the EAT

The computational model is not only a means to implement an emotion structure into the EAT according to appraisal theories of emotion, but also the main pedagogical mechanism through which meaning-making is meant to be extrapolated and harnessed by learners. The evidence gathered in the empirical contributions of the thesis sketches a contrasting picture with respect to the efficacy of the model, as highlighted in the data-driven assessment of Chapter \@ref(study-3). In this section I will therefore try to extend the analysis of the computational model in more general terms. First, I briefly resume the main indications that have been collected on the computational model and why they are problematic. Second, I provide and discuss some conceptual alternatives to the proposed model. Finally, I posit the reasons why the attempt to implement an appraisal-driven structure into the EAT should be carried on in spite of the empirical setbacks of the present contribution.

### Is It Worth Having a Computational Model Determined by Appraisal Theories of Emotion?

Data collected through the empirical contributions of the thesis have provided an ambivalent assessment of the computational model at the core of the proposed EAT. On the one hand, the computational model based on the Valence x Control/Power appraisals obtained a *fair* predictive accuracy as an expert system [@schererHumanEmotionExperiences2013; @moorsAppraisalTheoriesEmotion2013; @fontaineLinearNonlinearRelationships2021]. Most of the subjective feeling expressed belonged to the subset of computed suggestions, hinting that the propositions, even if far from a perfect match, are sufficiently accurate to communicate one's own subjective feeling. On the other hand, the rating on the two appraisal dimensions was consistent with previous work in highlighting a major overlapping between the Valence and the Control/Power dimensions, which tend to co-vary. As previous scholars have pointed out, it is safe to assume that this is mainly due to the overarching importance given to the Valence dimension, with the Control/Power following suit, conditioned on whether the situation is perceived as pleasant or unpleasant [@lavoueEmotionalDataCollection2017; @erbasRoleValenceFocus2015; @schererSemanticStructureEmotion2018].

This is clearly a major shortcoming both from a user experience and pedagogical perspective. For the former, the learner is asked for an explicit action that seems to add little independent value. The action is thus redundant and in contrast with the aim of parsimony and relevance in prompting user efforts [@rogersInterctionDesignHumanComputer2011; @mackenzieHumancomputerInteractionEmpirical2013]. For the latter, learners seem to gain little understanding from the Control/Value reflection, even though, at a theoretical level, this is a major determinant in discriminating emotional experience, as well as a relevant pedagogical dimension in influencing learning processes and outcomes [@shumanConceptsStructuresEmotions2014; @pekrunControlValueTheoryAchievement2006; @pekrunControlValueTheoryAchievement2014].

In Chapter \@ref(study-3) I have already suggested some possible solutions to the problem. I think that the possibility to exclude a *pure* Valence dimension may be the most promising one, forcing learners to evaluate the situation according to a different perspective [@erbasRoleValenceFocus2015]. The extension of the computational model to multi-dimensional affective spaces illustrated in Chapter \@ref(computational-model), unfortunately not tested in the present contribution, is also an attempt in the direction of a more accurate differentiation in the subjective feeling, which is consistent with the evidence that two dimensions are not enough for the whole gamut of emotional experiences [@gilliozMappingEmotionTerms2016; @scherer2018; @fontaineLinearNonlinearRelationships2021]. For instance, there may be cognitive evaluations of the situation at hand that are more or less relevant depending on characteristics of the learning context. This allows researchers and practitioners to target more specifically emotions linked to learning, as it is the case of epistemic emotions, which have only recently been put under the radar [@pekrunMeasuringEmotionsEpistemic2016; @shumanConceptsStructuresEmotions2014; @erdemliRoleEpistemicEmotions2021].

In the meantime, the solutions I have proposed are all of a *more-of-the-same* nature, whereas the results obtained warrant a deeper analysis of the model. In other words, it is legitimate to ask whether the presence of the computational model driven by appraisal theories of emotion really adds something to the EAT in terms of facilitating emotional awareness. The answer to this question can be broken down in two parts.

The first part concerns whether a computational model is useful in the first place, regardless of the inner workings. In this sense, even considering the mixed results obtained in the present contribution, I think that the computational perspective should be maintained. I share the view that computational models force scholars to be transparent about a phenomenon of interest. As stated by Guest and Martin [-@guestHowComputationalModeling2021, p. 790], "computational modeling [forces] scientists to explicitly document an instance of what their theory assumes, if not what their theory is".

The second part wonders whether the computational model should be informed by an appraisal driven-theory of emotion in the way the proposed model does. In order to answer this second part of the question, I will first briefly discuss some alternatives in the next subsection.

### Alternative Computational Models Based on An Active Evaluation fo the Situation

The way the interface of the proposed EAT is built makes the underlying computational model independent from the appearance of the tool. This paves the way to alternative computational models that can be implemented on the very same user interface. I identify three possible kind of computational models: theory-driven, data-driven, and scripting-driven. These are illustrated with the same conventions adopted in Chapter \@ref(computational-model):

-   $E$ represents the expected subjective feeling;

-   $A$ represents the set of appraisal criteria against which the situation is assessed.

The theory-driven approach is the one adopted in the current contribution, and is represented by the following equation:

$$
E = A + Theory
$$

In this approach, the likelihood of an expected subjective feeling is derived only by the theoretical model, represented by the affective space. In other words, the model is not influenced neither by contextual factors, nor by the information coming in from the use of the tool through time -- for instance, the previous emotion experienced by the learner. This way of considering an emotion is consistent with the idea, particularly relevant in appraisal-driven theories, of an emotion as a *Swiss-knife* that maximizes its usefulness in many different circumstances, depending on the person making the evaluation of which *tool* is more appropriate to the situation at hand on a moment-by-moment basis [@moorsAppraisalTheoriesEmotion2013; @scherer2019; @schererEmotionProcessEvent2019].

A first alternative to a pure theory-driven approach consists in harnessing the concrete use of the tool by the generated data, with the possibility to add the information about the previous emotion experienced. In this case, the computational model would not be an *expert system* with a fixed set of responses, but would adapt depending on the use that is made of the tool. With a platform built around the EAT, it would be possible to set up some form of machine learning based on collected data as a training set. The previous experienced emotion could also be adopted as a predictor, leveraging on the chances that, in learning contexts, there are transitions between emotions that are more likely to occur than others [@dmelloDynamicsAffectiveStates2012; @dmelloConfusionCanBe2014; @lehmanConfusionComplexLearning2012]. This approach does not exclude the possibility to inject theoretical knowledge, for instance as a form of Bayesian prior that regularizes the model [@mcelreathStatisticalRethinkingBayesian2020] or as a placeholder while a sufficient accuracy of the data-driven model is reached. An example of the application of a data-driven principle is provided by @arguedasModelProvidingEmotion2018, even though the system analyzes emotional cues in the text produced by students in an online environment rather than active self-reporting. The equation for a data-driven model would therefore be the following:

$$
E = A + {Data} + {Previous}
$$

The data-driven approach would therefore imply some form of *folk interpretation* of emotion, which, as already mentioned, I think is less interesting from a pedagogical perspective . As @molinariEMORELOutilReporting2016 have pointed out, learners could assign the same appraisal profile to very different subjective feelings. If a model was trained on this kind of data, the result could be highly unpredictable. This is even more relevant when considering that a data-driven perspective presupposes a sort of *black box* in the underlying algorithm, which, regardless of its accuracy, would nevertheless be opaque in scientific terms. This is of course a subject that transcends the specifics of this contribution and relates more broadly to the complex interplay between explanation and prediction [@schererEmotionProcessEvent2019; @schererPredictionDataDriven2019]. Still, a transparent computational model would be easier to critique, discuss, and adapt according to scientific inputs [@guestHowComputationalModeling2021]. The use of previous data is also ambivalent. On the one hand, it has the potential to situate emotion in the evolution of the learning task, providing a continuity between what has happened at $T_{-1}$, what is happening at $T_{0}$, and what may happen at $T_{1}$. At the same time, as illustrated in Chapter \@\ref(defining-emotion-unit), one of the adaptive function attributed to emotion is precisely to provide abrupt and unforseen discontinuity with respect to the situation at hand in what may be considered a Markovian process [@sanderModelsEmotionAffective2013; @scherer2019; @moorsAppraisalTheoriesEmotion2013]. The use of previous data could also potentially reinforce dysfunctional emotion dispositions due to appraisal biases [@scherer2021].

Another alternative consists in applying probabilities of experienced emotions given the instructional design that is planned for the task at hand. This is a common pedagogical strategy that may induce, for instance, a state of confusion to be resolved through appropriate pedagogical scripting [@dmelloConfusionCanBe2014; @lehmanConfusionComplexLearning2012; @voglSurpriseCuriosityConfusion2019]. Probabilities would therefore be weighted, for instance, depending on the timing in a series of learning sequences: in sequence 1 learners are more likely to experience $E$ and therefore should appraise the situation with $A$; whereas in sequence 2 the targeted experience is $E'$, which should trigger $A'$. This kind of model is disruptive compared to the previous ones, since there is a form of inverse probability at play. In the theory driven and data-driven approaches, we are interested in the probability of experiencing a certain emotion given the evaluation at hand, that is, $P({subjective \space feeling}|evaluation)$. On the contrary, with a scripting strategy, we are interested in inducing a cognitive evaluation that is driven by a targeted experience, for that particular emotion to be experienced by learners -- that is, $P(evaluation|{subjective \space feeling})$. In other words, the aim is to lead learners to realize that they have a particular evaluation of the situation that is congruent with the expected subjective feeling aimed by the learning design. In this case, thus, the computational model from a scripting perspective would consists in the following equation:

$$
A = E + Scripting
$$

The scripting-driven perspective provides a huge potential in pedagogical terms. It may in fact contribute to increase learners' emotional competence, particularly in the sense of an *appropriate appraisal* of the situation [@schererComponentialEmotionTheory2007; @lavoueEmotionAwarenessTools2020]. In a planned unfolding of emotional experiences, a scripting-driven algorithm may guide learners to the discovery of socio-cognitive tensions through an immediate and familiar lens -- that of prototypical emotions that may emerge according to the specifics of the situation at hand. If things go as planned, this would pave the way to more controlled and identifiable regulation processes, which may even be automated as in some affect aware systems discussed in Section \@ref(affect-aware-systems). On the other hand, if one assumes that emotions are dynamic processes, entangled with the variability of higher order cognitive functions, they cannot easily be defined in a linear and deterministic way, unless the event under scrutiny is highly predictable, as in the case of predefined stimuli. The experimental contribution of \@ref(study-1) illustrated to what extent the emotional experience may vary even when using a dummy task and a controlled environment. One can imagine that the variability in emotion experience is a function of the complexity of the learning setting: the more complex it is, the more variability may be expected. Furthermore, even if a scripting strategy would be transparent with respect to the underlying computational model, since the probability of $E$ could be formally parametrized, the script itself will be highly dependable on the task at hand. As a consequence, the model itself could hardly be generalized to other situations, unless two pedagogical scripts share almost exactly the same context, since changes in context may even switch the meaning of the same lexicalized emotion as @gentschEffectsAchievementContexts2017 have outlined.

### Some Tenacity May Be Temporarily Warranted

It follows from the discussion of the alternative above that there is not, to my mind, a conceptual computational model presenting a clear advantage over an appraisal-driven theoretical model. In spite of the limitations revealed in the actual use of the EAT in the present contribution, I consider premature to abandon it all together. Some form of "theoretical tenacity" [@meehlAppraisingAmendingTheories1990; @eronenTheoryCrisisPsychology2021] to the overall idea of implementing an appraisal-driven structure manifestly into the tool may therefore be warranted, if only from a *faute de mieux* perspective. I consider two main arguments in favor of this claim: the first is taken from advances about an overarching theory of emotion, whereas the second is more pragmatic in nature.

In a recent discussion on the topic of emotion science, Scherer [-@scherer2022] argues that times are ripe for theory convergence. The author analyzes how a componential model of emotion, as the one depicted in \@ref(fig:tf-cpm-general-model-adapted), provides considerable overlapping between competing emotion theories, limiting divergences to more detailed or conventional issues.

> In most cases, the existing differences seem to be due to the use of different terminology (e.g. construction, basic, discrete, affect, experience), on the respective level of analysis (e.g. brain, behaviour, language), and, in particular, on unequal emphasis placed on specific components and stages of the emotion process.\
> --- @scherer2022, p. 164

In his analysis of the commonalities, Scherer (*ibid.*) argues that all major emotion theories subsume, more or less explicitly, a direct link between (1) an ongoing evaluation of events, behaviors in oneself and others, situational factors, surrounding objects, and personal or collective memories; and (2) a partial access to the whole emotion process in the form of a feeling that can be labelled depending on personal subjectivity and collective conventions. As a consequence, an overt computational model as the one proposed here may therefore simply make explicit what, in other settings, is implied.

At the same time, the link between the appraisal and the subjective feeling components is neither the only, nor necessarily the more relevant one. For instance, one of the disadvantages of using the subjective feeling as part of the underlying structure is that consequences on behavior may have to be inferred. One can imagine to replace the subjective feeling with some representation of an action tendency, which is more overtly aimed at the consequences on behavior. In terms of the interface, this may result, for example, in buttons labelled with actions such as *I want to quit*, *I need more information*, or *I require assistance*, which are still driven by appraisal criteria. In other words, before abandoning the overall idea of the model, it is possible to act also outside the realm of the appraisal criteria as I have so far envisioned.

The second reason, of a more pragmatic nature, relates to the parsimony, or simplicity, of the model at hand. When models are concerned, the saying attributed to George Box [-@boxScienceStatistics1976] is often evoked: "all models are wrong, but some are useful". I think this should be interpreted as a contracted form of a more nuanced sentence: *all models don't capture the whole of the phenomena, but some of them are useful, to some extent and under specific assumptions*. In an inter-disciplinary field such as the investigation of emotional awareness in computer-mediated learning environments, disposing of a *simple* and transparent model -- wrapped up in a configurable tool -- may contribute to draw the attention about the need to specify assumptions about the affective side of the equation [@pekrunIntroductionEmotionsEducation2014; @dukesRiseAffectivism2021; @linnenbrink-garciaAdaptiveMotivationEmotion2016]. The transparency and simplicity of the model may represent a gentle introduction for researchers outside the domain of affective science, who nevertheless reckon the importance and interest of affective phenomena. In this regard, the model should also be weighted depending on the number of researchers and practitioners who see a potential utility for it, to an extent sufficient for its adoption and under the assumptions that are specific to their settings and goals (a topic that will be more thoroughly addressed in the conclusion). There is always the possibility, at a later and more defined stage of the process, to switch towards more complex and accurate computational models [@marsellaComputationalModelsEmotion2010; @schererPredictionDataDriven2019], as well as more articulated affect- or emotion-awares systems [@harleyDevelopingEmotionAwareAdvanced2017; @dmelloAffectiveComputingApproach2018].

### Synthesis

## Suggestions for the Implementation of Emotion Awareness Tools

-   Affective phenomenon or phenomena targeted

-   Why a theory in pedagogical terms

-   Different versions of the tool

-   Provide a description of lexicalized emotion and justification for the choice

-   Consider training learners before using

-   Provide measure of usability/ux experience

The proposed EAT is only one way of implementing a theory-driven structure of emotion into a tool. I consider competition in science a necessity for progress, and therefore I highly encourage other scholars to continue the development of alternative tools or embark in the development of new ones. To this extent, I provide some suggestions that may be of interest for such an enterprise, often by discussing other suggestions made by other scholars.

1.  The tool should not only be anchored to an emotion theory, but the link should also be illustrated in terms of what are the advantages for learners to implement that theory rather than another.

2.  The tool should make explicit which kind of affective phenomena it targets

3.  The tool should justify the kind and number of dimensional/discrete emotion unit among which learners can choose

4.  
