---
bibliography: references.bib
---

# (PART) Concluding Remarks {.unnumbered}

# General Discussion {#general-discussion}

```{r setup-general-discussion}
library(dagitty)
```

The general discussion of the thesis draws both from the implementation of the proof of concept in Part II and the results of the empirical contribution in Part III. Since the previous chapters have already proposed extensive analyses of the specific topics or research questions at hand, I will adopt a higher-level perspective for the general discussion, focusing first on what I think are the two main contributions of the thesis: the abstract model of emotional awareness, and the computational model at the core of the proposed EAT. For each model, I propose an analysis of its contributions and limitations. I then use the overall experience of implementing and assessing an EAT to provide what I think are useful suggestions for other scholars interested in providing emotional awareness through a dedicated tool. The chapter ends with future perspectives, which are organized on the theoretical, technical, and empirical levels.

## Analysis of the Abstract Model of Emotional Awareness

One of the main objectives of the thesis was related to the construct validity of emotional awareness as a concept. I reckon that the construct as so far been defined mostly implicitly following what I think is a mild form of *psychologisme*: given that emotion play a prominent role in everyday life and it is also related to mechanisms that are implicated to learning, then it is safe to deduce that it should also play a prominent role in computer-mediated learning environments. In a recent and forming field of research, some amount of implicit assumptions is not only necessary, but also instrumental in attempting to determine where research should be directed [@fiedlerCycleTheoryFormation2004]. To further clarify the concept in an attempt to improve construct validity, I have proposed an abstract model, based on three underlying assumptions of emotional awareness: first, that emotion play a prominent role at the inter-personal level; second, that emotion convey useful information at the inter-personal level; and third, that emotion can be fruitfully expressed and perused in computer-mediated learning environments.

(ref:thesis-scm-model-repeated-caption) Abstract model of the functions of an EAT from the perspective of a single learner disposing of it in a computer-mediated learning environment. Repeated from Figure @ref(fig:thesis-scm-model).

```{r thesis-scm-model-repeated}
#| fig.cap="(ref:thesis-scm-model-repeated-caption)", 
#| fig.align= "center",
#| fig.scap = "Abstract model of emotional awareness", 
#| out.width= "100%"
knitr::include_graphics(
    here::here("figure/intro/thesis-scm-model.png")
)
```

The abstract model, illustrated again in Figure \@ref(fig:thesis-scm-model-repeated), consists in reducing emotional awareness to a core of simpler functions, for which some potential intervening factors are provided. It starts from the learning activity as a source of potential emotional experiences and paves two parallel paths at the intra-personal and inter-personal levels, culminating in meaning-making, which can be related back to the learning activity. In this section, I propose a critical evaluation of the model by integrating support -- or lack thereof -- of empirical data.

### Disentangling Intra- and Inter-Personal Emotional Awareness

The model has, to my sense, the merit of uncovering some of the implicit assumptions that are made in considering emotional awareness instrumental to computer-mediated learning environments. In particular, it draws a clearer distinction between the intra-personal and inter-personal perspectives, going thus in the direction of "disentangling the effect of reflecting upon one's own emotions from the effect of sharing one's emotions with the partner" [@molinariEmotionFeedbackComputermediated2013, p.342]. Highlighting more clearly the intra- and inter-personal building blocks of emotional awareness paves the way to the possibility of configuring tools and environments in which the different forms of emotional awareness are more or less explicitly sought after and enhanced.

The experiment of Chapter \@ref(strudy-1), in which the use of and access to emotional information was manipulated, is a first attempt in this direction. Even if the results have been undermined by the small sample size at hand, the experiment at least succeeded in pointing out how, in the definition of emotional awareness as "information about one's own emotions and/or about the emotions of other learners", the *and/or* distinction has profound consequences on the mechanisms at play in producing and perusing emotional awareness. In this regard, I think that it is important to have some performance-based metrics, which allows to quantify the actual use of the tool. It is known that, especially in high-demanding tasks, users of a technological artifact can have an altered perception of what they have been really doing with it. Especially with a dedicated tool that falls outside the content space of a learning task, actual indications are particularly revealing. To this extent, the measurement of eye-tracking transitions between zones of the computer-mediated learning environment seem to me to be the most promising indicator of the integration between emotional information and the learning process.

In the longitudinal study of Chapter \@ref(study-2), the distinction between the two paths was less clear-cut, assured only by the three word clouds used to represent the affiliation and distancing function of emotion identified by Fischer and Manstead [-@fischerSocialFunctionsEmotion2016]. Nevertheless, some empirical data also allowed to reveal the two paths. For instance, the evaluation of the Emotion Awareness Usefulness (EAU) pointed out how learners had higher expectations on the inter-personal dimensions compared to the intra-personal dimensions. The distinction on the scale was further corroborated by the presence, in the exploratory factor analysis, of a self-centered cluster (*Self-Understanding* and *Self-Regulation dimensions*) and a partner(s) oriented cluster (*Social Presence*, *Understanding Others*, and *Self-Other Comparison* dimensions). In other words, the distinction between the intra- and inter-personal paths is not only pertinent in the concrete, technical functioning of the EAT, but also in the perceived usefulness.

The overall empirical data collected in both studies seem to privilege an inter-personal stance of the EAT. Even though this may seem natural, since we are comparing an information that is already known by the learner (intra-personal) with something that is not (inter-personal), in a self-report context the two levels can easily become entangled by the fact that intra-personal awareness is conditioned to the quality of intra-personal awareness. Furthermore, there may also be a sort of *altruistic effect*: by thinking that one is mainly producing emotional awareness for others to peruse, one is actually benefiting from producing the data in the first place.

Unfortunately, the thesis hasn't tested a *pure* intra-personal setting, where there is no contact whatsoever with other learners. In the *Self-Centered* condition of Study 1, there was still some form of communication and social presence with another learner, even though a simulated one. Moreover, the participant knew -- or suspected at least -- that the other person was also reflecting on her emotions, since this was an overt request for the empirical task. In Study 2, even if the setting was individual on the learning tasks, the aim was for emotional awareness to contribute to an overall sense of *socio-affective* presence and sense of belonging.

It would therefore be interesting to replicate the use of an EAT similar to that of the study by Molinari and colleagues [-@molinariEMORELOutilReporting2016] with the EMORE-L tool, depicted in Chapter \@ref(eat-general-chapter), but using the proposed EAT with the possibility to express emotions on a moment-by-moment basis. In Molinari and colleagues' study (*ibid.*), the authors highlighted how students praised for the possibility to dispose of inter-personal emotional awareness. In a direct comparison of the use of a similar EAT between a *pure* individual setting and more socially oriented conditions, it would be possible to quantify the different use and perception of usefulness of intra- vs. inter-personal emotional awareness.

### The Link Between Meaning-Making and Learning Remains Mostly Implicit

On the downside, the abstract model falls short in purporting the left-hand side of the model: how meaning-making is extrapolated and related back to the learning activity. Even though it was not an explicit objective of the thesis to explore this relationship above and beyond the role of an EAT, this mechanism is at the core of why emotional awareness should be provided in the first place. The thesis made the overt assumption that, for this passage even to occur, it is important to maximize the chances that meaning-making is made on solid ground from an *affective-first* perspective.

I reckon that this assumption is open to valid criticism: whether the ground is solid should also be determined from a pedagogical perspective. A mainstream position on this subject is that learners can be guided by instructional designs or adaptive learning environments towards emotional experiences that are inherently instrumental to learning. I have chosen to take an alternative path consistent with the awareness tool perspective: learners are bestowed with the locus of control in considering when and how emotional information can contribute to a better (re)assessment of the learning situation ad hand, leveraging on meaning-making that is extrapolated from the emotional information available.

As highlighted in the theoretical part of the thesis, this assumption requires a series of events to fall into place, which are difficult to assess before one is assured about if and how emotional awareness can be fruitfully conveyed in the first place. That is the main reason why the thesis has only scraped the surface of this pivotal underlying assumption and rather focused on the overall features, use and perception of the EAT at hand. In doing so, though, I reckon that I have in somehow neglected the potential role of the EAT in giving to emotional awareness a pedagogical dimension. The longitudinal study in particular has confirmed that, without an adequate pedagogical scenario in which the EAT is inserted, it is also a form of *psychologisme* to expect that endowing learners with emotional awareness would miraculously create the conditions for a socio-emotional dimension to emerge.

The pedagogical role for extrapolating meaning-making was almost totally deferred to the emotion structure injected into the tool, notably by the computational model at the core of the EAT. In the previous chapters, I have focused on the analysis of the model mainly from an UX perspective, even in the data-driven assessment of Chapter \@ref(study-3). I now reckon that the computational model plays a more prominent role of what I had anticipated also in the left-hand side of the abstract model. From a pedagogical perspective, it assists learners in making and reinforcing inferences about the link between the evaluation of the situation at hand and what they may probably experience at the conscious emotional level. In the next section I therefore propose a more holistic analysis of the computational model.

## Analysis of the Computational Model at the Core of the EAT

As stated in the previous section, the computational model is not only the way through which an emotion structure is implemented according to appraisal theories of emotion, but also the main pedagogical mechanisms through which meaning-making is meant to be extrapolated and harnessed by learners in the situation at hand. The evidence gathered in the empirical contributions of the thesis sketches a contrasting picture with respect to the efficacy of the model, as highlighted in the data-driven assessment of Chapter \@ref(study-3). On the one hand, the computational model based on the Valence x Control/Power appraisals obtained a *good-enough* predictive accuracy as an expert system. Most of the subjective feeling expressed belonged to the subset of computed suggestions, hinting at the fact that the proposition are sufficiently accurate to communicate one's own subjective feeling. On the other hand, the rating on the two appraisal dimensions confirmed previous work in highlighting a major overlapping between the Valence and the Control/Power dimensions, which tend to co-vary. As previous scholars have pointed out, it is safe to assume that this is mainly due to the overarching importance given to the Valence dimension, with the Control/Power following suit, conditioned on whether the situation is perceived as pleasant or unpleasant.

This is clearly a major shortcoming both from a user experience and pedagogical perspective. For the former, the learner is asked for an explicit action that seems to add little independent value. As a consequence, the action is redundant and in contrast with the general outlook in user experience advocating parsimony and relevance in prompting user action. For the latter, learners seem to gain little understanding from the Control/Value reflection, even though, at a theoretical level, this is a major determinant in discriminating emotional experience, and also a relevant pedagogical dimension in influencing learning processes and outcomes.

The extension of the computational model to multi-dimensional affective spaces illustrated in Chapter \@ref(computational-model) was mainly driven by this major issue. Allowing researchers and practitioners to adopt the number and kind of appraisal dimensions, which also determine the likelihood of the subjective feeling, is a concrete attempt to overcome the issue. Depending on the specifics of learning activity and the computer-mediated learning environment, there may be cognitive evaluations of the situation at hand that are more or less relevant. This leaves the door open to future work targeting more specifically emotions linked to learning, as it is the case, for instance, to epistemic emotions, which have only recently been put under the radar.

In this section, I will nevertheless evaluate the computational model in more general terms, and compare it with potential alternative based on the same core principle: there is an active evaluation of the situation at hand, according to which potential subjective experiences receive different weights. The next section provides a brief outlook of two alternatives, and the following discuss why I consider the theory-driven approach to still be the most viable candidate besides the highlighted shortcomings.

### Alternative Computational Models Based on An Active Evaluation fo the Situation

As already mentioned, the way the interface of the proposed EAT is built makes the underlying computational model independent from the appearance of the tool. This paves the way to alternative computational models that can be implemented on the very same user interface. I identify three possible kind of computational models: theory-driven, data-driven, and scripting-driven. These are illustrated with the same conventions adopted in illustrating the computational model in Chapter \@ref(computational-model):

-   $E$ represents the expected subjective feeling, and

-   $A$ represents the set of appraisal criteria against which the situation is assessed.

The theory-driven approach is the one adopted in the current contribution, and is represented by the following equation:

$$
E = A + Theory
$$

In this approach, the likelihood of an expected subjective feeling is derived only by the theoretical model, represented by the affective space. In other words, the model is not influenced neither by contextual factors, nor by the information coming in from the use of the tool through time -- for instance, the previous emotion experienced by the learner. This way of considering an emotion is consistent with the idea, particularly relevant in appraisal-driven theories, of an emotion as a *Swiss-knife* that maximizes its usefulness in many different circumstances, depending on the person making the evaluation of which *tool* is more appropriate to the situation at hand on a moment-by-moment basis.

A first alternative to pure theory-driven approach consists in harnessing the concrete use of the tool by the generated data, with the possibility to add the information about the previous emotion experienced. The equation for a data-driven model would therefore be the following:

$$
E = A + {Data} + {Previous}
$$

In this case, the computational model would not be an *expert system* with a fixed set of responses, but would adapt depending on the use that is made of the tool. With a platform built around the EAT, it would be possible to set up some form of machine learning that uses existing data as training set. The previous experienced emotion could also be adopted as a predictor, leveraging on the chances that, in learning contexts , there are transitions between emotions that are more likely to occur than others [@dmelloDynamicsAffectiveStates2012; @dmelloConfusionCanBe2014; @lehmanConfusionComplexLearning2012]. This approach does not exclude the possibility to inject theoretical knowledge, for instance as a form of Bayesian prior that regularizes the model [@mcelreathStatisticalRethinkingBayesian2020].

Another alternative consists in applying probabilities of experienced emotions given the instructional design that is planned for the task at hand. This is a common pedagogical strategy that may induce, for instance, a state of confusion, which may be resolved through appropriate pedagogical scripting. Probabilities would therefore be weighted, for instance, depending on the timing in a series of learning sequences. In this case, the computational model from a scripting perspective would consists in the following equation:

$$
E = A + Scripting
$$

This kind of model is disruptive compared to the previous ones, since there is a form of inverse probability at play. In the theory driven and data-driven approaches, we are interested in the probability of experiencing a certain emotion given the evaluation at hand, that is, $P(emotion|evaluation)$. On the contrary, with a scripting strategy, we are interested in inducing a cognitive evaluation that is driven by a targeted emotion, for that particular emotion to be experienced by the learners, that is, $P(evaluation|target \space emotion)$.

### Theory-Driven Remains the Better Candidate

Even though the theory driven approach has shown limitations in the current contribution, I still think it is the most viable option, especially from a multi-purpose perspective.

The data-driven approach would imply some form of *folk interpretation* of emotion, which, as already mentioned, I think is less interesting from a pedagogical perspective. I maintain the opinion that an EAT should rather explicitly assume a *prescriptive* interpretation of what an emotion is and what is it for.

The pedagogy-driven perspective, on the other hand, does not really provide learners with the means to use emotion as valuable information: what is sought after is rather the induction of an affective experience, regardless of learners' ongoing evaluation, based on the assumption that it contributes to a learning process. I stick to the idea that emotions are dynamic processes, entangled with higher order cognitive functions, which cannot easily be defined in a linear and deterministic way, unless the event under scrutiny is highly predictable, as in the case of a video lecture. Otherwise, I consider the variability in emotion experience as a function of the complexity of the learning setting: the more complex it is, the more variability is to be expected. This variability can be better accounted for when a probabilistic link between the evaluation of the situation and the overall experience is accounted by an overarching emotion theory.

All things considered, thus, I posit that the implementation of a theory-driven EAT should be continued. Especially in a parsimonious form, it may be a relatively affordable tool to enhance emotional awareness in different computer-mediated learning environments.

## Suggestions for the Implementation of Emotion Awareness Tools

The proposed EAT is only one way of implementing a theory-driven structure of emotion into a tool. I consider competition in science a necessity for progress, and therefore I highly encourage other scholars to continue the development of alternative tools or embark in the development of new ones. To this extent, I provide some suggestions that may be of interest for such an enterprise, often by discussing other suggestions made by other scholars.

1.  The tool should not only be anchored to an emotion theory, but the link should also be illustrated in terms of what are the advantages for learners to implement that theory rather than another.

2.  The tool should make explicit which kind of affective phenomena it targets

3.  The tool should justify the kind and number of dimensional/discrete emotion unit among which learners can choose

4.  

## Future Perspectives

### From an Abstract to a Structural Causal Model of Emotional Awareness

-   Brief introduction to SCM and DAG

-   Exemple of controlling for gender in CSCL experiments

-   Attempt to draw a preliminary SCM

### Technical Improvements to the EAT

### Further Empirical Adoption of the EAT
