---
bibliography: references.bib
---

# Conclusion {.unnumbered}

```{r conclusion-setup, include=FALSE, echo=FALSE}
library(tidyverse) 
library(papaja) 
library(see) 
library(RColorBrewer)
library(here) 
library(knitr) 
library(kableExtra)

Sys.setenv(LANG = "en")

# Load the relevant data and graphs from the study-3 folder

source(here("data", "generative-model", "simulation-of-participant-emotions.R"), local = knitr::knit_global(), encoding = "UTF-8")
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE) 
```

In this contribution I have outlined the implementation and assessment of an Emotion Awareness Tool (EAT) based on the assumption that learners may benefit from emotional awareness in computer-mediated learning environment if they (1) reflect on their emotion from a subjective feeling experience rather than autonomic recognition; (2) guide their reflection on the appraisal of a situation and convey emotional awareness to others using the same information unity they have used to extrapolate meaning-making from the situation at hand; and (3) dispose of emotional awareness on a moment-to-moment basis, which bestow to learners the responsibility of producing and perusing emotional information whenever they see fit.

The theory-driven implementation and the empirical assessment have mutually influenced each other resulting in the following main contributions:

1.  Following the need to better clarify the concept, an abstract model of emotional awareness in computer-mediated learning environment was first proposed as means to outline common functions of Emotion Awareness Tools more broadly. The abstract model was then transformed through the lenses of the overall contribution into a Structural Causal Model of the effect of emotional awareness on learning processes and outcomes.
2.  A computational model that originally linked only two appraisal dimensions to a subjective feeling according to an underlying bidimensional affective space has been extended to an *N*-dimensional structure in an attempt to best comply with a greater number of appraisals that can contribute to emotion elicitation and differentiation. The computational principle can also be applied more widely to a componential approach to emotion as long as a form of computation of emotional responses (action tendencies, motor expressions or subjective feelings) is derived from a cognitive evaluation of the situation at hand.
3.  A toolbox is proposed to researchers and practitioners interested in endowing computer-mediated learning environments with emotional awareness. The toolbox tries to foster transparency and comparability of results in research by providing the possibility to configure the tool. The configuration itself can be leveraged as an intervention/independent variable, whose effect can be measured on phenomena of interest.
4.  In this regard, two types of measures have been proposed as useful to the overall investigation and assessment of the use of an Emotion Awareness Tool: (1) the use of gaze transitions between the tool and part of the screen where the learning task takes place, and (2) an Emotion Awareness Usefulness scale, which, to the best of my knowledge, represents a first attempt to provide a common measure to the many facets implied in the use of an EAT.

The thesis adopted an overall perspective aiming primarily at the construct and internal validities of emotional awareness as a concept. Such an attempt stretches over a wide domain of inter-dependent and inter-disciplinary aspects. In an individual contribution, especially at the doctoral stage, the work thus inevitably highlights a number of limitations or uncompleted steps in the process. This concluding chapter starts by pointing out three main limitations and then provides a glimpse of three different perspectives that may be taken for future developments. Finally, some thoughts about the overall work are proposed by the author.

## Limitations

Three main limitations are identified in the present contribution. The first relates to the absence of elements that are more prominently related to learning. The second concerns the lack of empirical data, which would have fostered a more severe assessment of the tool. Finally, despite the interaction and user-centered approach adopted by the thesis, few researchers or practitioners have been implicated in the process.

### Where Is Learning in All This?

In spite of the overall context referring to computer-mediated learning environment, there is a clear imbalance between the *affective* and the *learning* facets of the thesis. Even though the focus on emotional awareness was anticipated and justified by the willingness to work on construct and internal validity, one may argue that those very features cannot be separated from the overall goal of improving learning processes and outcomes. In other words, emotional awareness as intended in the use of an EAT exists only insofar as its benefits can be directly related to learning.

The neglect of the learning side can be identified with two main reasons. First, in the few contributions that have investigated the effect of emotional awareness on learning processes and outcomes, there is a great variability both in how emotional awareness is implemented and in which *facets* of learning were targeted: problem-solving, quality of interaction, understanding of the partner, emotion regulation, etc. [@avrySharingEmotionsContributes2020; @avryAchievementAppraisalsEmotions2020; @avryExplicitSharingEmotions2021; @eligioEmotionUnderstandingPerformance2012; @molinariEmotionFeedbackComputermediated2013; @feidakisProvidingEmotionAwareness2014; @molinariEMORELOutilReporting2016; @lavoueEmotionalDataCollection2017; @lavoueEmotionAwarenessTools2020]. Second, it is hard to determine a time and learning design scale in which emotional awareness can have an effect. Once again, previous studies (*ibid.*) vary greatly on the matter, ranging from relatively short tasks in a controlled environment to ecological implementation in existing courses.

The thesis attempted to deal with these two problems by holding the emotional awareness side more or less constant and vary the learning part as much as possible following the objective to provide a multipurpose tool. As a result, the learning side was inevitably undermined by too great a variability of settings, tasks, and participants' profiles. Nonetheless, these sort of problems are common to most scientific endeavors aiming at the investigation of learning. Thus, they cannot represent, by themselves, a justification. There was certainly the means to do more.

For instance, in the Emotional Awareness Usefulness scale introduced in Chapter \@ref(study-2), there is not a single item that is explicitly related to the contribution of emotional awareness to learning processes and outcomes. It is auto-referenced towards emotional awareness itself. In a scale that aims at measuring the usefulness of emotional awareness in computer-mediated learning environments, one may expect to assess, among other things, whether and how the presence of emotional awareness contributed to the overall learning processes and outcomes. The scale should therefore be extended beyond the affective realm.

Another aspect overlooked in the thesis are epistemic emotions, which are tightly related to learning [@muisRoleEpistemicEmotions2018; @erdemliRoleEpistemicEmotions2021; @audrinInterestConfusionContrasting2022; @chevrierExploringAntecedentsConsequences2019; @voglSurpriseCuriosityConfusion2019; @muisCuriousCaseClimate2015]. Having adopted the overarching Component Process Model as the main theoretical framework, emotion has been conceptualized in abstract and generalized terms for any potential situation at hand [@schererStudyingAppraisaldrivenEmotion2019; @schererDynamicArchitectureEmotion2009; @schererWhatAreEmotions2005]. One may consistently argue that in a computer-mediated learning environments, the intrinsic nature of the stimuli to which the organism/learner is confronted are only a subset of *real life*. This has consequences both on the appraisal dimensions that may be more revealing to adopt, as well as the lexicalized emotions representing the subjective feelings of learners [@shumanConceptsStructuresEmotions2014]. There is a growing literature that considers epistemic emotions from an appraisal/componential perspective [@audrinInterestConfusionContrasting2022; @chevrierExploringAntecedentsConsequences2019; @voglSurpriseCuriosityConfusion2019]. Their integration would therefore be facilitated even holding the same characteristics of the EAT.

To sum up, the subtitle of the thesis *with an application in computer-mediated learning environments* is quite unsatisfactory. It is certainly true in terms of situational factors, but it lacks depths and explicit links in conceptual terms.

### Contribution is Mainly Analytic

A second main shortcoming of the thesis concerns the lack of empirical data. The first study yielded interesting data on the use of the tool, but from a limited sample size. Conversely, the ecological setting of the second study was promising both in sample size and timescale, but fell short in the use of the tool. As a consequence, the overall claims advanced by the thesis could not really be assessed by a severe test [@mayoStatisticalInferenceSevere2018; @meehlAppraisingAmendingTheories1990; @anvariNotAllEffects2022; @scheelWhyHypothesisTesters2021]. The contribution remains mainly analytic in nature, with the features and prospected benefits of the tool derived by combining affective science principles with assumptions inferred from the relatively scant literature on the role of emotion in computer-mediated learning environments. This limitation has both conceptual and practical reasons.

From the conceptual standpoint, the thesis started with a perspective that was more tightly related to the EATMINT project presented in Chapter \@ref(dew-chapter). The scope was therefore mainly the influence of emotion on Computer-Supported Collaborative Learning [@dillenbourgSymmetryPartnerModelling2016; @molinariEmotionFeedbackComputermediated2013; @cereghettiSharingEmotionsComputermediated2015; @avrySharingEmotionsContributes2020], as corroborated by the first study. As already mentioned, though, I progressively started to realize that the way emotional awareness was implemented should be determined before testing specific hypothesis [@scheelWhyHypothesisTesters2021]. So the focus of the thesis drifted towards the implementation and assessment of the EAT.

On the practical side, this shift was incidentally reinforced by the global pandemic: a third study ready to be carried out was first stopped by the lock-down and then adapted to be conducted online. Finally, it was postponed for in the meantime I had been focusing on the implementation of the EAT from a broader perspective. Since one of the aim of pursuing a thesis in psychology was to reorient myself from a former web development and UX background, this limitation of the thesis is the one that is personally more deceiving.

### Few Researchers Implicated So Far

The third main limitation of the thesis concerns the small number of researchers that have adopted or contributed to the development of the tool so far. As a consequence, the features of the EAT are once again mainly derived by reinterpreting needs and potential adoptions from the extant literature, or even more bluntly by the specific needs of the thesis.

This shortcoming is to a great extent due to my personal habit to work in isolation, which I have more recently came to realize as dysfunctional even from a scientific perspective. I nowadays value the importance of collaborative projects, even at large scales [@moshontzPsychologicalScienceAccelerator2018; @stroebeWhatCanWe2019]. To a lesser extent, though, the shortcoming was also determined by the fact that researchers who manifested their interest for the tool were mainly looking for two types of solutions, none of which was really provided by the proposed EAT.

The first type of solution relates to the need of a *simple* digitalized self-report tool [@mortillaroEmotionsMethodsAssessment2015]. In other words, researchers were not interested in the moment-to-moment, multi-user, or data-persistence features. On the contrary, they rather sought the possibility to express an emotion as an induced response to a predetermined stimulus.

The second type of solution refers to *ad hoc* tools, which are tightly integrated into the specific learning environment, as is the case of affect- or emotion-aware systems presented in Chapter \@ref(general-affective-background). Researchers were actually looking for the possibility not only to create emotional awareness, but also to dispose of an environment which adapted or reacted depending on learners' emotions.

In both cases researchers were mostly disappointed rather than intrigued by the main philosophy and features of the toolbox. I nevertheless think it is too soon to determine if this is the sign of an intrinsic and stable shortcoming. I hope that with more fruitful examples of its use, researchers and practitioners may be encouraged to extend beyond the two dominant perspectives. Yet, one potential and ironic limitation of the thesis could still be that, by aiming for a tool that may be adopted by the many, it actually fell short in meeting anybody's needs!

## Future Perspectives

Three main areas of future development are proposed. The first concerns a wider involvement of interested scholars in the further development of the toolbox. The second refers to the possibility to extend the use of the computational model as a generative algorithm. Finally, a potential line of inquire is identified in the empirical comparison between a scripting and an awareness strategy to provide learners with emotional awareness.

### Usability Test of the Toolbox

A first future step, which is aimed specifically as a remedy to the third limitation outlined above, consists in conducting a usability test [@lallemandMethodesDesignUX2018; @tullisMeasuringUserExperience2013] of the toolbox with researchers and practitioners from different backgrounds. The presence of specific tasks and scenarios to accomplish may contribute to give a glimpse about the possibilities of the tool, leaving at the same time the necessary flexibility to consider specific needs not yet covered.

The usability test should take into account at least two potential *persona*. On the one hand, researchers that are primarily interested in the affect-related consequences of the use of the tool [@cerneaSurveyTechnologiesRise2015; @lavoueEmotionAwarenessTools2020; @lavoueEmotionalDataCollection2017]. For them, setting up an appropriate affective space and disposing of accurate measures would probably be the most important features [@mortillaroEmotionsMethodsAssessment2015; @ritchieEvolutionSelfreportingMethods2016; @silvaComparativeStudyUsers2020]. On the other hand, researchers or practitioners that are primarily interested in learning processes and outcomes. This second type of *persona* is particularly meaningful considering how the learning realm has been neglected so far. In this case, the focus should be more about the relational and content spaces of instructional design [@janssenCoordinatedComputerSupportedCollaborative2013; @kirschnerAwarenessCognitiveSocial2015; @jarvelaCapturingDynamicCyclical2019; @dillenbourgSymmetryPartnerModelling2016].

### Generative Use of the Computational Model

From a computational standpoint, a future perspective concerns the use of the computational model as a generative algorithm [@mcelreathStatisticalRethinkingBayesian2020; @marsellaComputationalModelsEmotion2010; @guestHowComputationalModeling2021]. In fact, since the computational model works with any number of values on appraisal dimensions and any affective space as *database*, the algorithm can be adopted to generate emotions on the basis of specific, identifiable and transparent criteria.

For instance, in the aforementioned empirical contribution that was planned but not carried out, the experimental plan consisted in pairing participants with four different simulated partners. The simulated partners differed according to the emotional experience they manifested during the same joint problem-solving task as in Chapter \@ref(study-1). The four resulting conditions would have been determined by *emotional profiles* characterized by a combination of high vs. low *Valence* and high vs. low *Control/Power*. A similar manipulation was adopted by @avrySharingEmotionsContributes2020, even though only from a dimensional perspective. To build the four different *emotional profiles* the following process was considered:

1.  Take the mean and standard deviation of the *Valence* and *Control/Power* appraisals from a meta-analysis of previous contributions adopting the same task;
2.  Add or subtract three quarters of the standard deviation on the mean of each appraisal dimension to obtain a manipulated mean for the higher or lower modality;
3.  Draw a number of appraisal evaluations consistent with the number of emotions to simulate from a normal distribution determined by the two computed means and the pooled standard deviation from the meta-analysis;
4.  Inject the values of the two appraisal dimensions for each emotion into the computational model and retrieve a subjective feeling that is consistent with the appraisal profile. This could be done, for instance, by adding some form of sampling within the proposed subset to avoid the prevalence of the same lexicalized emotions.

Figure \@ref(fig:simulated-emotions-manipulation) illustrates the result of applying the computational model to generate 26 simulated emotions. For each emotion, the two appraisal values and the lexicalized emotions are provided. The four normal distributions from which the appraisal values were drawn are derived from empirical data in Fritz (2015) and from Chapter \@ref(study-1). By varying the distance between the high and low values on each dimension, the emotion profiles can be manipulated to be more or less extreme. It may also be possible to adopt a time series to have a more nuanced fluctuation.

(ref:simulated-emotions-manipulation-caption) Use of the computational model as a generative algorithm in order to simulate 26 emotions from normal distributions defined by the mean and standard deviation of previous empirical data.

```{r simulated-emotions-manipulation}
#| out.width="100%",
#| fig.height=7,
#| fig.cap="(ref:simulated-emotions-manipulation-caption)",
#| fig.scap="Use of the computational model as a generative algorithm"
gm.simulated_emotions.graph
```

Considering that the *Valence* x *Control/Power* dimensions have outlined serious overlapping, it is maybe for the better that this manipulation was eventually not carried out. But the process could still be valid with another underlying affective space. The advantage is that a simulated partner could be formally determined, so that the emotional experience can be theoretically justified, but also computationally replicated.

### Scripting vs. Awareness Strategy

Finally, from the empirical point of view, an interesting further step would be to investigate the moment-to-moment awareness feature of the EAT in comparison to a scripting strategy [@millerScriptingAwarenessTools2015; @buderGroupAwareness2021]. Even though I have advocated the interest to explore how different settings of the same EAT may perform rather than the with-or-without emotional awareness approach [@buderGroupAwarenessTools2011], it is still a major concern whether ongoing emotional awareness is a strength or a nuisance.

An interesting middle ground would therefore consists in comparing two different ways of providing emotional information. One in which the information is always available and bestow to learners the responsibility to produce and peruse that information. The other in which the instructional design impose to learners moments in the task for considering the emotional experience. To increase the comparability of the two settings, one can imagine to adopt, in the scripting condition, a strategy similar to @tranEmotionsDecisionMakingProcesses2012, who used the Geneva Emotion Wheel [@schererGRIDMeetsWheel2013; @schererWhatAreEmotions2005] to create emotional awareness in collaborative teams at specific moments (see Chapter \@ref(dew-chapter) for more details). In this way, the emotional structure would be mostly the same in the two conditions, with learners extrapolating awareness both from the appraisal and subjective feeling components.

Given that the efficacy of each strategy may also depend on the characteristics of the task at hand (difficulty, interdependence with other colleagues, etc.), the type of task may represent a second independent variable. This would be a more severe test of the contribution of an EAT, providing useful evidence on the claim that the dynamic and ongoing nature of emotion should be transposed in a computer-mediated learning task.

## Final Thoughts

> It was the best of times, it was the worst of times, it was the age of wisdom, it was the age of foolishness, it was the epoch of belief, it was the epoch of incredulity, it was the season of Light, it was the season of Darkness, it was the spring of hope, it was the winter of despair, we had everything before us, we had nothing before us, we were all going direct to Heaven, we were all going direct the other way.\
> --- Charles Dickens, A Tale of Two Cities

I started my academic career a few months after the Open Science Collaboration project published its report *Estimating the reproducibility of psychological science* [@opensciencecollaborationEstimatingReproducibilityPsychological2015]. At the time, I was mostly unaware of all the problems underlying psychology as a science. Throughout the thesis I had to face the different *crisis* that were outlined and the methodological solutions that were proposed. It was a source both of great anxiety and excitement. On the one hand, it seemed that every step in any direction could be undermined by theoretical or empirical shortcomings. As a side effect, the most conservative position would have been immobilism. On the other hand, it was also a great opportunity to explore new approaches and question current practices. The result is inevitably a scattered mosaic of old and new: a process rather than a contribution *stricto sensu*.

The thesis also took place during the pandemic, with millions of students swiped into distance learning overnight. With my lab, we produced in a hurry some basic resources to share with university teachers without any previous experience outside *ex-cathedra* teaching [@tecfaEnseignerDistanceDans2019]. As a result, while for the thesis I was working on how socio-affective aspects of learning should be taken into account, for the actual reality of higher education I was basically telling people "*no, you can't just record a two-hour lecture and expect students to watch it alone at home*". Once again, this was also a source of mixed feelings. On the one hand, it was legitimate to doubt the practical use of the thesis given how far it was from reality. On the other hand, I also witnessed students struggling, helping each other out with the means they had, and achieving their learning goals in the face of adversity. If any of this can represent even the tiniest help in similar -- but hopefully less dramatic -- situations, then it was certainly worth the while.
