---
bibliography: references.bib
---

# Conclusion {.unnumbered}

```{r conclusion-setup, include=FALSE, echo=FALSE}
library(tidyverse) 
library(papaja) 
library(see) 
library(RColorBrewer)
library(here) 
library(knitr) 
library(kableExtra)

Sys.setenv(LANG = "en")

# Load the relevant data and graphs from the study-3 folder

source(here("data", "generative-model", "simulation-of-participant-emotions.R"), local = knitr::knit_global(), encoding = "UTF-8")
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE) 
```

I posited that the whole concept of emotional awareness could be better grasped if learners (1) reflect on their emotion from a subjective feeling experience rather than autonomic recognition; (2) guide their reflection on the appraisal of a situation and convey emotional awareness to others using the same information unity they have used to extrapolate meaning-making from the situation at hand; and (3) dispose of emotional awareness on a moment-by-moment basis, which bestow to learners the responsibility of producing and perusing emotional information at the moment the see fit.

## Limitations

### Where Is Learning in All This?

-   Studies vary in emotional awareness and learning tasks, attempt to hold one of them constant

-   Difficulty to find a middle ground for representation of learning (small tasks or all courses)

-   

### Contribution is Mainly Analytic

-   Started mainly in CSCL context following previous work

-   Planned experience could not be carried out

-   Focus on the Toolbox

### Few Researchers Implicated So Far

-   Big error from user-centered design

-   Researchers often seek :

    -   a custom implementation of "simple" data collection

    -   a tool that is tightly integrated with the learning environment

-   Is there really an interest for a decoupled tool ?

## Future Perspectives

### Usability Test of the Toolbox

-   Usability test, Focus groups, ...

-   Work on graphical representations

### Generative Use of the Computational Model

```{r simulated-emotions-manipuation}
#| out.width="100%",
#| fig.height=7,
#| fig.cap="Simulated emotions from the computational model"
gm.simulated_emotions.graph
```

### Scripting vs. Awareness strategy
