---
bibliography: references.bib
---

# Conclusion {.unnumbered}

```{r conclusion-setup, include=FALSE, echo=FALSE}
library(tidyverse) 
library(papaja) 
library(see) 
library(RColorBrewer)
library(here) 
library(knitr) 
library(kableExtra)

Sys.setenv(LANG = "en")

# Load the relevant data and graphs from the study-3 folder

source(here("data", "generative-model", "simulation-of-participant-emotions.R"), local = knitr::knit_global(), encoding = "UTF-8")
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE) 
```

In this contribution I have outlined the implementation and assessment of en Emotion Awareness Tool based on the assumption that learners may benefit from emotional awareness in computer-mediated learning environment if they (1) reflect on their emotion from a subjective feeling experience rather than autonomic recognition; (2) guide their reflection on the appraisal of a situation and convey emotional awareness to others using the same information unity they have used to extrapolate meaning-making from the situation at hand; and (3) dispose of emotional awareness on a moment-by-moment basis, which bestow to learners the responsibility of producing and perusing emotional information whenever they see fit.

The theory-driven implementation and the empirical assessment have mutually influenced each other resulting in the following main contributions:

1.  Following the need to better clarify the concept, an abstract model of emotional awareness in computer-mediated learning environment has been proposed. The model has then been declined more specifically in an appraisal-driven perspective, in which the evaluation of the situation plays a central role in a Structural Causal Model of the effect of emotional awareness on learning processes and outcomes.
2.  A computational model that originally linked only two appraisal dimensions to a subjective feeling according to an underlying bidimensional affective space has been extended to an *N*-dimensional structure in an attempt to best comply with a greater number of appraisals that can contribute to emotion elicitation and differentiation. The computational principle can also be applied more widely to a componential approach to emotion as long as a form of computation of emotional responses (action tendencies, motor expressions or subjective feelings) is derived from a cognitive evaluation of the situation at hand.
3.  A toolbox is proposed to researchers and practitioners interested in endowing computer-mediated learning environments with moment-to-moment emotional awareness. The toolbox tries to foster transparency and comparability of results in research by providing the possibility to configure the tool. The configuration itself can be leverage as an intervention/independent variable, whose effect can be measured on phenomena of interest.
4.  In this regard, two types of measures have been proposed as useful to the overall investigation and assessment of the use of an Emotion Awareness Tool: (1) the use of gaze transitions between the tool and part of the screen when the learning task takes place, and (2) an Emotion Awareness Usefulness scale, which, to the best of my knowledge, represents a first attempt to provide a common measure to the many facets implied in the use of an Emotion Awareness Tool in computer-mediated learning environments.

The thesis adopted an overall perspective aiming more broadly to the construct and internal validity of emotional awareness. Such an attempt, especially in an individual contribution at the doctoral stage, stretches over a wide domain of inter-dependent and inter-disciplinary aspects. As a consequence, the work certainly highlights a number of limitations or uncompleted steps in the process. This concluding chapter thus starts by pointing out some of the main limitations of the present work. It then provides a glimpse for future perspective. Finally, some thoughts about the overall work are proposed by the author.

## Limitations

### Where Is Learning in All This?

-   Studies vary in emotional awareness and learning tasks, attempt to hold one of them constant

-   Difficulty to find a middle ground for representation of learning (small tasks or all courses)

-   Epistemic emotion overlooked

### Contribution is Mainly Analytic

-   Started mainly in CSCL context following previous work

-   Planned experience could not be carried out

-   Focus on the Toolbox

### Few Researchers Implicated So Far

-   Big error from user-centered design

-   Researchers often seek :

    -   a custom implementation of "simple" data collection

    -   a tool that is tightly integrated with the learning environment

-   Is there really an interest for a decoupled tool ?

## Future Perspectives

### Usability Test of the Toolbox

-   Usability test, Focus groups, ...

-   Work on graphical representations

### Generative Use of the Computational Model

```{r simulated-emotions-manipuation}
#| out.width="100%",
#| fig.height=7,
#| fig.cap="Simulated emotions from the computational model"
gm.simulated_emotions.graph
```

### Scripting vs. Awareness strategy
