---
bibliography: references.bib
---

# Conclusion {.unnumbered}

```{r conclusion-setup, include=FALSE, echo=FALSE}
library(tidyverse) 
library(papaja) 
library(see) 
library(RColorBrewer)
library(here) 
library(knitr) 
library(kableExtra)

Sys.setenv(LANG = "en")

# Load the relevant data and graphs from the study-3 folder

source(here("data", "generative-model", "simulation-of-participant-emotions.R"), local = knitr::knit_global(), encoding = "UTF-8")
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE) 
```

In this contribution I have outlined the implementation and assessment of en Emotion Awareness Tool based on the assumption that learners may benefit from emotional awareness in computer-mediated learning environment if they (1) reflect on their emotion from a subjective feeling experience rather than autonomic recognition; (2) guide their reflection on the appraisal of a situation and convey emotional awareness to others using the same information unity they have used to extrapolate meaning-making from the situation at hand; and (3) dispose of emotional awareness on a moment-by-moment basis, which bestow to learners the responsibility of producing and perusing emotional information whenever they see fit.

The theory-driven implementation and the empirical assessment have mutually influenced each other resulting in the following main contributions:

1.  Following the need to better clarify the concept, an abstract model of emotional awareness in computer-mediated learning environment has been proposed. The model has then been declined more specifically in an appraisal-driven perspective, in which the evaluation of the situation plays a central role in a Structural Causal Model of the effect of emotional awareness on learning processes and outcomes.
2.  A computational model that originally linked only two appraisal dimensions to a subjective feeling according to an underlying bidimensional affective space has been extended to an *N*-dimensional structure in an attempt to best comply with a greater number of appraisals that can contribute to emotion elicitation and differentiation. The computational principle can also be applied more widely to a componential approach to emotion as long as a form of computation of emotional responses (action tendencies, motor expressions or subjective feelings) is derived from a cognitive evaluation of the situation at hand.
3.  A toolbox is proposed to researchers and practitioners interested in endowing computer-mediated learning environments with moment-to-moment emotional awareness. The toolbox tries to foster transparency and comparability of results in research by providing the possibility to configure the tool. The configuration itself can be leverage as an intervention/independent variable, whose effect can be measured on phenomena of interest.
4.  In this regard, two types of measures have been proposed as useful to the overall investigation and assessment of the use of an Emotion Awareness Tool: (1) the use of gaze transitions between the tool and part of the screen when the learning task takes place, and (2) an Emotion Awareness Usefulness scale, which, to the best of my knowledge, represents a first attempt to provide a common measure to the many facets implied in the use of an Emotion Awareness Tool in computer-mediated learning environments.

The thesis adopted an overall perspective aiming more broadly to the construct and internal validity of emotional awareness. Such an attempt, especially in an individual contribution at the doctoral stage, stretches over a wide domain of inter-dependent and inter-disciplinary aspects. As a consequence, the work certainly highlights a number of limitations or uncompleted steps in the process. This concluding chapter thus starts by pointing out some of the main limitations of the present work. It then provides a glimpse for future perspective. Finally, some thoughts about the overall work are proposed by the author.

## Limitations

### Where Is Learning in All This?

In spite of the overall context of the thesis referring to computer-mediated learning environment, there is a clear imbalance between the *affective* and the *learning* sides of the thesis. Even though the focus on emotional awareness was anticipated and justified by the willingness to work on construct and internal validity of the concept, one may argue that those very features cannot be separated from the overall goal of improving learning processes and outcomes. In other words, emotional awareness as intended in the use of an emotional awareness tool exists only insofar as its benefit can be directly related to learning. The neglect of the learning side can be related to two main reasons.

First, in the few contributions that have investigated the effect of emotional awareness on learning processes and outcomes, there is a great variability both in how emotional awareness is implemented and in which *facets* of the complex compound of *learning* were targeted: problem-solving, quality of interaction, understanding of the partner, emotion regulation, etc. [@avrySharingEmotionsContributes2020; @avryAchievementAppraisalsEmotions2020; @avryExplicitSharingEmotions2021; @eligioEmotionUnderstandingPerformance2012; @molinariEmotionFeedbackComputermediated2013; @feidakisProvidingEmotionAwareness2014; @molinariEMORELOutilReporting2016; @lavoueEmotionalDataCollection2017; @lavoueEmotionAwarenessTools2020].

Second, it is hard to determine a time and learning design scale in which emotional awareness can have an effect. One again, previous studies (*ibid.*) vary greatly on the matter, ranging from relatively short tasks in a controlled environment to ecological implementation in existing courses.

The thesis attempted to deal with these two problems by holding the emotional awareness side more or less constant and vary the learning part as much as possible following the objective to provide a multipurpose tool. But considering that these two problems are common to most scientific endeavors related to the investigation of learning, they cannot be, by themselves, a justification. There was certainly the possibility to do more.

For instance, in the Emotional Awareness Usefulness scale there is not a single item that is explicitly related to the contribution of emotional awareness to learning processes and outcomes. It is auto-referenced towards emotional awareness itself. In a scale that aims at measuring the usefulness of emotional awareness in computer-mediated learning environments, one may expect to also assess whether and how the presence of emotional awareness contributed to the overall learning processes and outcomes. The scale should therefore be extended beyond the affective realm.

Another aspect overlooked by the overall thesis are epistemic emotions, which are tightly related to learning [@muisRoleEpistemicEmotions2018; @erdemliRoleEpistemicEmotions2021; @audrinInterestConfusionContrasting2022; @chevrierExploringAntecedentsConsequences2019; @voglSurpriseCuriosityConfusion2019; @muisCuriousCaseClimate2015]. Having adopted the overarching Component Process Model as the main theoretical framework, emotion has been conceptualized in abstract and generalized terms for any potential situation at hand [@scherer2019; @schererDynamicArchitectureEmotion2009a; @schererWhatAreEmotions2005]. One may consistently argue that in a computer-mediated learning environment, the intrinsic nature of the stimuli to which the organism/learner is confronted are only a subset of *real life*. This has consequences both on the appraisal dimensions that may be more revealing to adopt, as well as the lexicalized emotions representing the subjective feelings of learners. There is a growing literature that considers epistemic emotions from an appraisal/componential perspective [@audrinInterestConfusionContrasting2022; @chevrierExploringAntecedentsConsequences2019; @voglSurpriseCuriosityConfusion2019]. Their integration would therefore be facilitated even holding the same characteristics of the emotion awareness tool.

To sum up, the subtitle *with application to computer-computer learning environment* stated in the title of the thesis is quite unsatisfactory. It is certainly through in terms of situational factors, but it lacks depths and explicit links on conceptual terms.

### Contribution is Mainly Analytic

A second main shortcoming of the thesis concerns the lack of empirical data. The first study yielded interesting data on the use of the tool, but from a limited sample size. Conversely, the second study's ecological setting was promising in sample size and duration, but fell short in the use of the tool. As a consequence, the overall claims advanced by the thesis could not really be assessed by a severe test [@mayoStatisticalInferenceSevere2018; @meehlAppraisingAmendingTheories1990; @anvariNotAllEffects2022; @scheel2020]. The contribution remains thus mainly analytic in nature, with the features and prospected benefits of the tool derived by combining affective science principles with assumptions inferred from the relatively scant literature on the role of emotion in computer-mediated learning environments. This limitation has both conceptual and practical reasons.

From the conceptual standpoint, the thesis started with a perspective that was more tightly related to the EATMINT project presented in Chapter \@ref(dew-chapter). The scope was therefore mainly the influence of emotion on Computer-Supported Collaborative Learning [@dillenbourgSymmetryPartnerModelling2016; @molinariEmotionFeedbackComputermediated2013; @cereghettiSharingEmotionsComputermediated2015; @avrySharingEmotionsContributes2020], as it is corroborated for instance by the first study. As already mentioned, though, I then started to realize that the way emotional awareness was implemented should be determined before testing specific hypothesis [@scheel2020]. So the focus of the thesis progressively drifted towards the implementation and assessment of the EAT.

On the practical side, this shift was incidentally reinforced by the pandemic: a third study that was ready to be carried out was first stopped by the lock-down, then adapted to be conducted online, and finally postponed for in the meantime I had been focusing on the implementation of the EAT from a broader perspective.

Since one of the aim of pursuing a thesis in psychology was to stop developing (which was my previous profession) and start collecting and analyzing empirical data, this limitation of the thesis is the one that is personally more deceiving.

### Few Researchers Implicated So Far

The third main limitations of the thesis relates to the limited number of researchers that have adopted or contributed to the development of the tool. In spite of the user-centered approach that has been adopted throughout the thesis, few researchers were directly implicated in the process. As a consequences, the features of the EAT are once again mainly derived by the specific needs of the thesis or by reinterpreting needs and potential adoptions from the extant literature.

This shortcoming is to a great extent due to my personal habit to work in isolation: an habit that I fortunately came to realize to be wrong even from a scientific perspective. I nowadays value the importance of projects implicating many researchers from many different countries and backgrounds [@moshontzPsychologicalScienceAccelerator2018; @stroebeWhatCanWe2019].

To a lesser extent, though, the shortcoming was also determined by the fact that researchers who manifested interested for the tool were mainly looking for two types of solutions, which are different from the emotion awareness tool proposed by the thesis. The first type of solution relates to the need of a *simple* digitalized self-report tool [@mortillaroEmotionsMethodsAssessment2015]. In other words, researchers were not interested in the moment-to-moment and multi-user perspective of the tool, but rather sought the possibility to express an emotion as an induced response to a predetermined stimulus. The second type of solution refers to an ad-hoc tool that is tightly integrated with the learning environment, as is the case for affect-aware systems presented in Chapter \@ref(general-affective-background). In both cases, thus, researchers were mostly disappointed rather than intrigued by the features of the toolbox.

I think it is too soon to determine whether this is the consequences of an intrinsic and stable limitation of the tool, or it is rather due by the lack of fruitful empirical examples that may encourage researchers or practitioners to switch from the two dominant perspectives. Still, one potential ironic limitation of the thesis is that by aiming for a tool that may be adopted by the many, it fall shorts of meeting anyone's needs.

## Future Perspectives

### Usability Test of the Toolbox

A first future step, which is aimed specifically as a remedy to the third limitation outlined above, consists in conducting a usability test [@lallemandMethodesDesignUX2017; @tullisMeasuringUserExperience2013] of the toolbox with researchers and practitioners from different backgrounds. The presence of specific tasks and scenarios to accomplish may contribute to give a glimpse to researchers about the possibilities of the tool, leaving at the same time the necessary flexibility to consider their specific needs.

The test should take into account at least two potential *persona*. On the one hand, researchers that are primarily interested in the affect-related consequences of the use of the tool [@cerneaSurveyTechnologiesRise2015; @lavoueEmotionAwarenessTools2020; @lavoueEmotionalDataCollection2017]. For them, setting up an appropriate affective space and disposing of interesting measures would probably be the most important features [@scherer2018; @mortillaroEmotionsMethodsAssessment2015; @ritchieEvolutionSelfreportingMethods2016; @silvaComparativeStudyUsers2020]. On the other hand, researchers or practitioners that are primarily interested in learning processes and outcomes. This second type of *persona* would be particularly meaningful considering how the learning realm has been neglected so far. In this case, the focus should be more about the relational and content spaces of instructional design [@janssenCoordinatedComputerSupportedCollaborative2013; @kirschnerAwarenessCognitiveSocial2015; @jarvelaCapturingDynamicCyclical2019; @dillenbourgSymmetryPartnerModelling2016].

### Generative Use of the Computational Model

```{r simulated-emotions-manipuation}
#| out.width="100%",
#| fig.height=7,
#| fig.cap="Simulated emotions from the computational model"
gm.simulated_emotions.graph
```

### Scripting vs. Awareness strategy
